<script>`r readAllFile(system.file("js", "utils-report.js", package = "patRoon"))`</script>

Annotation {data-orientation=rows}
===

##

### feature groups { data-width=300 .fGroups }

<style> .fGroups { overflow-x: auto; } </style>

```{r echo=FALSE}
compFGroups <- compsClustFGroups <- formFGroups <- componentFGroups <- mfWebFGroups <- character()
if (!is.null(rmdVars$compounds))
{
    compTable <- compoundTable(rmdVars$compounds)
    compFGroups <- rmdVars$groupNames[sapply(rmdVars$groupNames,
                                             function(grp) !is.null(compTable[[grp]]) && nrow(compTable[[grp]]) > 0)]
}
if (!is.null(rmdVars$compsCluster))
{
    cl <- clusters(rmdVars$compsCluster)
    compsClustFGroups <- rmdVars$groupNames[rmdVars$groupNames %in% names(cl)]
}
if (rmdVars$includeMFWebLinks != "none")
{
    if (rmdVars$includeMFWebLinks == "compounds")
        mfWebFGroups <- compFGroups
    else if (rmdVars$includeMFWebLinks == "MSMS")
        mfWebFGroups <- rmdVars$groupNames[sapply(rmdVars$groupNames,
                                                  function(grp) any(sapply(peakLists(rmdVars$MSPeakLists),
                                                                           function(pa) !is.null(pa[[grp]]) &&
                                                                               !is.null(pa[[grp]][["MSMS"]]))),
                                                  USE.NAMES = FALSE)]
    mfWebLinks <- sapply(mfWebFGroups, function(grp)
    {
        if (grp %in% compFGroups && is(rmdVars$compounds, "compoundsMF"))
        {
            # make link with used MF settings
            set <- settings(compounds)
        }
        else
            set <- NULL
        
        return(buildMFLandingURL(set, rmdVars$MSPeakLists[[grp]][["MSMS"]],
                                 rmdVars$gInfo[grp, "mzs"]))
    })
}
if (!is.null(rmdVars$formulas) && "formulas" %in% rmdVars$reportPlots)
{
    formFGroups <- groupNames(rmdVars$formulas)
    formFGroups <- intersect(formFGroups, rmdVars$groupNames)
}
if (!is.null(rmdVars$components))
{
    cTable <- componentTable(rmdVars$components)
    componentFGroups <- unique(unlist(sapply(cTable, "[[", "group")))
    componentFGroups <- componentFGroups[componentFGroups %in% rmdVars$groupNames]
}
plotGroups <- unique(c(compFGroups, compsClustFGroups, formFGroups))
allGroups <- unique(c(plotGroups, componentFGroups, mfWebFGroups))

# UNDONE: replace by proper inheritance
isSusp <- isScreening(rmdVars$fGroups)
keepSuspCols <- character()
colsRound5 <- "mz"
if (isSusp)
{
    fGroupsSpecDT <- as.data.table(rmdVars$fGroups, collapseSuspects = NULL)
    
    isSet <- isFGSet(rmdVars$fGroups)
    keepSuspCols <- c("susp_name", "susp_d_rt", "susp_d_mz")

    clR2 <- c("susp_d_rt", "susp_annSimForm", "susp_annSimComp", "susp_annSimBoth", "susp_maxFragMatchesRel")
    if (isSet)
        clR2 <- getAllSuspSetCols(clR2, names(fGroupsSpecDT), sets(rmdVars$fGroups))
    fGroupsSpecDT[, (clR2) := lapply(mget(clR2), round, 2)]

    colsRound5 <- c(colsRound5, "susp_d_mz")

    # NOTE: all cols must be same type
    mergeCols <- function(fg, curColNames, parColNames, mergedName, is = isSet)
    {
        if (is)
        {
            for (s in sets(rmdVars$fGroups))
                fg <- mergeCols(fg, paste0(curColNames, "-", s), parColNames, paste0(mergedName, "-", s), FALSE)
            return(fg)
        }
        
        present <- which(curColNames %in% names(fg))
        if (length(present) == 0)
            return(fg)
        
        newColName <- paste(mergedName, paste0("(", paste0(parColNames[present], collapse = "/"), ")"))
        keepSuspCols <<- c(keepSuspCols, newColName)
        
        fg[, (newColName) := do.call(paste, c(mget(curColNames[present]), list(sep = " / ")))]
        return(fg)
    }

    fGroupsSpecDT <- mergeCols(fGroupsSpecDT, c("susp_formRank", "susp_compRank"), c("form", "comp"), "rank")
    fGroupsSpecDT <- mergeCols(fGroupsSpecDT, c("susp_annSimForm", "susp_annSimComp", "susp_annSimBoth"),
                               c("form", "comp", "both"), "annotated sim")
    fGroupsSpecDT <- mergeCols(fGroupsSpecDT, c("susp_maxFrags", "susp_maxFragMatches", "susp_maxFragMatchesRel"),
                               c("suspect", "max matches", "max matches rel"), "fragments")
    
    if (isSet)
        keepSuspCols <- c(keepSuspCols, "susp_sets", getAllSuspSetCols("susp_estIDLevel", names(fGroupsSpecDT),
                                                                       sets(rmdVars$fGroups)))
    else
        keepSuspCols <- c(keepSuspCols, "susp_estIDLevel")
} else
    fGroupsSpecDT <- as.data.table(rmdVars$fGroups)

fGroupsSpecDT <- fGroupsSpecDT[group %in% allGroups]
if (rmdVars$retMin)
    fGroupsSpecDT[, ret := ret / 60]

if (!is.null(fGroupsSpecDT[["neutralMass"]]))
    colsRound5 <- c(colsRound5, "neutralMass")

fGroupsSpecDT[, ret := round(ret, 2)]
fGroupsSpecDT[, (colsRound5) := lapply(mget(colsRound5), round, 5)]

fGroupsSpecDT <- fGroupsSpecDT[, intersect(c("group", "ret", "mz", "adduct", "neutralMass", keepSuspCols),
                                           names(fGroupsSpecDT)), with = FALSE]
fGroupsSpecDT[, groupInd := match(group, names(rmdVars$fGroups))]

showButton <- function(title, jsFunc, ...)
{
    args <- paste0("'", unlist(list(...)), "'", collapse = ", ")
    sprintf("<button onclick=\"%s(%s);\" style=\"padding: 0px 3px 0px 3px\">%s</button>", jsFunc, args, title)
}
maybeAddButton <- function(g, subGroups, ...) if (g %in% subGroups) showButton(...) else "^"

compButtons <- sapply(fGroupsSpecDT$group, function(g) maybeAddButton(g, compFGroups, "compounds", "showAnnotation",
                                                                      match(g, names(rmdVars$fGroups)), "compounds"))
compCLButtons <- sapply(fGroupsSpecDT$group, function(g) maybeAddButton(g, compsClustFGroups,
                                                                        "compounds clust", "showCompoundsCluster",
                                                                        match(g, names(rmdVars$fGroups))))
formButtons <- sapply(fGroupsSpecDT$group, function(g) maybeAddButton(g, formFGroups, "formulas",
                                                                      "showAnnotation", match(g, names(rmdVars$fGroups)),
                                                                      "formulas"))
mfWebButtons <- sapply(fGroupsSpecDT$group, function(g) maybeAddButton(g, mfWebFGroups, "MetFrag web",
                                                                       "window.open", mfWebLinks[g],
                                                                       "_blank"))

sp <- paste0(rep("&nbsp;", 4), collapse = "")
buttons <- paste(compButtons, compCLButtons, formButtons, mfWebButtons, sep = sp)
buttons <- gsub("\\^(&nbsp;)*", "", buttons) # remove placeholder (^) with accompanying spaces
fGroupsSpecDT[, show := buttons]
setcolorder(fGroupsSpecDT, c("groupInd", "group", "ret", "mz", "show"))

if (!is.null(rmdVars$components))
{
    annCols <- c("isogroup", "isonr", "charge", "ppm", # RAMClustR
                 "isotopes", "adnr", "adduct_rule", "adduct_charge", "adduct_nmol", "M_adduct", # CAMERA
                 "adduct_ion", # RC/CAMERA
                 "hsnr", "rGroup") # nontarget
    
    fGroupsSpecDT[, components := sapply(group, function(grp)
    {
        cmps <- findFGroup(rmdVars$components, grp)
        if (length(cmps) == 0)
            return("")
        
        return(wrapStr(paste0(sapply(cmps, function(cmpi)
        {
            cline <- cTable[[cmpi]][group == grp]
            if (nrow(cline) > 1) # might happen if there are multiple rGroups involved with NT components
            {
                cline[, rGroup := paste0(rGroup, collapse = "/")]
                cline <- cline[1]
            }
            cline <- cline[, sapply(cline, function(x) !is.na(x) && nzchar(x),
                                    USE.NAMES = FALSE), with = FALSE]
            annColsPresent <- annCols[annCols %in% names(cline)]
            cname <- names(rmdVars$components)[cmpi]
            
            if (length(annColsPresent) > 0)
            {
                cline <- cline[, annColsPresent, with = FALSE]
                for (j in seq_along(cline))
                {
                    if (is.numeric(cline[[j]]))
                        set(cline, 1L, j, round(cline[[j]], 5))
                }
                ann <- paste0(sprintf("%s: %s", names(cline), cline), collapse = ", ")
                return(sprintf("%s (%s)", cname, ann))
            }
            return(cname)
        }), collapse = ", "), width = 50, sep = "<br>"))
    }, USE.NAMES = FALSE)]
}

DT::datatable(fGroupsSpecDT,
              extensions = "Buttons",
              options = list(paging = FALSE, pageLength = -1, scrollX = TRUE, scrollY = "200px",
                             dom = "Bfrtip",
                             initComplete = DT::JS("function(settings, json)",
                                                   "{ setTimeout(initAnnotation, 25); }"),
                             order = list(list(1, "asc")),
                             columnDefs = list(list(visible = FALSE, targets = 0),
                                               list(className = "dt-nowrap", targets = 4),
                                               list(className = "dt-center",
                                                    targets = (seq_len(ncol(fGroupsSpecDT))[-5])-1)),
                             buttons = list(list(extend = "colvis", background = FALSE,
                                                 columns = seq(5, ncol(fGroupsSpecDT)-1),
                                                 collectionLayout = "three-column"))),
              escape = FALSE, rownames = FALSE, elementId = "fGroupsTable")
```

### EIC { data-width=100 }

<img id=EIC style="display:none;"></img>
<div id=noAnnotationSelected>Push a **show** button to view annotation data for a feature group.</div>


```{r allAnnPlots, fig.keep='none', eval=length(plotGroups) > 0}
# Generate all plots in advance, since having many code chunks will cause a lot of overhead.

message("Generating spectra...")
prog <- openProgBar(0, length(plotGroups))

plotPathFull <- getPlotPath(FALSE)
plotPathLink <- getPlotPath(TRUE)

allPlots <- setNames(lapply(seq_along(plotGroups), function(i)
{
    grp <- plotGroups[i]
    grpi <- match(grp, names(rmdVars$fGroups))
    grpPlots <- list()
    
    if (grp %in% compFGroups)
    {
        cTable <- compTable[[grp]]
        compsSeq <- seq_len(nrow(cTable))
        grpPlots[["compoundScores"]] <- sapply(compsSeq, function(compi)
        {
            ret <- file.path(plotPathFull, sprintf("compscore_%d_%d.png", grpi, compi))
            makeCachedPlot(ret, "plotScores", list(rmdVars$compounds, compi, grp, rmdVars$compoundsNormalizeScores,
                                                   rmdVars$compoundsExclNormScores, rmdVars$compoundsOnlyUsedScorings),
                           4.5, 4.5, bg = NA, cacheDB = rmdVars$cacheDB)
            return(ret)
        })
        
        grpPlots[["compoundSpectra"]] <- sapply(compsSeq, function(compi)
        {
            ret <- file.path(plotPathFull, sprintf("compspec_%d_%d.png", grpi, compi))
            makeCachedPlot(ret, "plotSpectrum", list(rmdVars$compounds, compi,  grp, rmdVars$MSPeakLists,
                                                     rmdVars$formulas, FALSE),
                           7, 4.5, bg = NA, cacheDB = rmdVars$cacheDB)
            return(ret)
        })
        
        grpPlots[["compoundStructs"]] <- sapply(compsSeq, function(compi)
        {
            ret <- file.path(plotPathFull, sprintf("compstruct_%d_%d.png", grpi, compi))
            makeCachedPlot(ret, "plotStructure", list(rmdVars$compounds, compi, grp, width = 150, height = 150),
                           3, 3, bg = NA, cacheDB = rmdVars$cacheDB)
            return(ret)
        })
    }
    
    if (grp %in% compsClustFGroups)
    {
        plotf <- file.path(plotPathFull, sprintf("dendro_%d.png", grpi))
        makeCachedPlot(plotf, "plot", list(rmdVars$compsCluster, groupName = grp), 8, 4.5, cacheDB = rmdVars$cacheDB)
        grpPlots[["compClustDendro"]] <- plotf
        
        ct <- cutClusters(rmdVars$compsCluster)[[grp]]
        grpPlots[["compClustMCS"]] <- sapply(seq_along(unique(ct)), function(cli)
        {
            ret <- file.path(plotPathFull, sprintf("mcs_%d_%d.png", grpi, cli))
            makeCachedPlot(ret, "plotStructure", list(rmdVars$compsCluster, grp, cli, 100, 100),
                           3, 3, cacheDB = rmdVars$cacheDB)
            return(ret)
        })
    }
    
    if (grp %in% formFGroups)
    {
        precursors <- unique(rmdVars$formulas[[grp]][["neutral_formula"]])
        
        grpPlots[["formulaScores"]] <- sapply(precursors, function(prec)
        {
            ret <- file.path(plotPathFull, sprintf("formscore_%d_%s.png", grpi, prec))
            makeCachedPlot(ret, "plotScores", list(rmdVars$formulas, prec, grp,
                                                   normalizeScores = rmdVars$formulasNormalizeScores,
                                                   excludeNormScores = rmdVars$formulasExclNormScores),
                           4.5, 4.5, bg = NA, cacheDB = rmdVars$cacheDB)
            return(ret)
        })
        
        grpPlots[["formulaSpecs"]] <- sapply(precursors, function(prec)
        {
            anPList <- annotatedPeakList(rmdVars$formulas, prec, grp,
                                         MSPeakLists = rmdVars$MSPeakLists, onlyAnnotated = TRUE)
            if (is.null(anPList))
                return("") # No MS/MS data available
            
            ret <- file.path(plotPathFull, sprintf("formspec_%d_%s.png", grpi, prec))
            makeCachedPlot(ret, "plotSpectrum", list(rmdVars$formulas, prec, grp, MSPeakLists = rmdVars$MSPeakLists),
                           6, 4.5, cacheDB = rmdVars$cacheDB)
            return(ret)
        })
    }
    
    setTxtProgressBar(prog, i)
    
    return(grpPlots)
}), plotGroups)
close(prog)

ap <- unlist(allPlots); ap <- ap[nzchar(ap)]
if (rmdVars$optimizePng && length(ap > 0))
    optimizePngPlots(ap)

if (rmdVars$selfContained)
    allPlots <- rapply(allPlots, function(ap) sapply(ap, function(p) if (nzchar(p)) knitr::image_uri(p) else ""), how = "replace")
```


## { .annotationClass .compounds }

### { .annotationClass .compounds }

<style> .compounds { overflow-x: auto; } </style>

```{r echo=FALSE, eval=length(compFGroups) > 0}
compoundsDT <- rbindlist(lapply(compFGroups, function(grp)
{
    mn <- mergedCompoundNames(rmdVars$compounds)

    ct <- compTable[[grp]]
    
    infoTexts <- sapply(seq_len(nrow(ct)), function(compi)
    {
        it <- paste0(getCompInfoList(ct, compi, TRUE, mn), collapse = "<br>")
        if (isSusp)
        {
            # insert suspect names (if any)
            tbl <- as.data.table(rmdVars$fGroups, collapseSuspects = NULL)[group == grp]
            if (!is.null(tbl[["susp_compRank"]]) && any(tbl$susp_compRank == compi, na.rm = TRUE))
                it <- paste(paste("<strong>Suspect(s):</strong>", paste0(tbl[susp_compRank == compi]$name, collapse = ", ")),
                             it, sep = "<br>")
        }
        return(it)
    })
    infoTexts <- sprintf("<div style='max-width: 300px; max-height: 432px; border: 1px solid black; border-style: dotted; margin: 1px; padding: 1px; overflow: auto; white-space: nowrap; }'>%s</div>", infoTexts)
    
    fiTables <- sapply(seq_len(nrow(ct)), function(compi)
    {
        apl <- annotatedPeakList(rmdVars$compounds, index = compi, groupName = grp,
                                 MSPeakLists = rmdVars$MSPeakLists, formulas = rmdVars$formulas,
                                 onlyAnnotated = TRUE)
        
        if (is.null(apl) || nrow(apl) == 0)
            return("<div align=\"center\">No annotation available.</div>")
        
        apl[, formula := subscriptFormulaHTML(formula)]
        apl[, neutral_loss := subscriptFormulaHTML(neutral_loss)]
        
        knitr::kable(apl, "html", escape = FALSE) %>%
                           kableExtra::kable_styling(font_size = 11) %>%
                           kableExtra::scroll_box(extra_css = "overflow: auto; height: 125px;")
    })
    
    return(data.table(group = match(grp, names(rmdVars$fGroups)),
                      "#" = seq_len(nrow(ct)),
                      compound = paste0(imgTags(allPlots[[grp]]$compoundStructs), "<br>", infoTexts),
                      spectrum = paste0(imgTags(allPlots[[grp]]$compoundSpectra), "<br>", fiTables),
                      scores = imgTags(allPlots[[grp]]$compoundScores)))
}))

DT::datatable(compoundsDT, options = list(scrollX = TRUE, scrollY = "600px", deferRender = TRUE,
                                          dom = "lrtp", pageLength = 25, autoWidth = TRUE,
                                          ordering = FALSE,
                                          columnDefs = list(list(visible = FALSE, targets = 0))),
              rownames = FALSE, escape = FALSE, elementId = "compoundsTable")
```


## { .annotationClass .formulas }

### { .annotationClass .formulas }

<style> .formulas { overflow-x: auto; } </style>

```{r echo=FALSE, eval=length(formFGroups) > 0}
formulasDT <- rbindlist(lapply(formFGroups, function(grp)
{
    precursors <- unique(rmdVars$formulas[[grp]][["neutral_formula"]])
    
    infoTexts <- sapply(seq_along(precursors), function(preci)
    {
        it <- paste0(getFormInfoList(rmdVars$formulas[[grp]], precursors[preci], useHTML = TRUE),
                     collapse = "<br>")
        if (isSusp)
        {
            # insert suspect names (if any)
            tbl <- as.data.table(rmdVars$fGroups, collapseSuspects = NULL)[group == grp]
            if (!is.null(tbl[["susp_formRank"]]) && any(tbl$susp_formRank == preci, na.rm = TRUE))
                it <- paste(paste("<strong>Suspect(s):</strong>", paste0(tbl[susp_formRank == preci]$name, collapse = ", ")),
                             it, sep = "<br>")
        }
        return(it)
    })
    infoTexts <- sprintf("<div style='max-width: 300px; max-height: 432px; border: 1px solid black; border-style: dotted; margin: 1px; padding: 1px; overflow: auto; white-space: nowrap; }'>%s</div>", infoTexts)
    
    fiTables <- sapply(precursors, function(prec)
    {
        apl <- annotatedPeakList(rmdVars$formulas, precursor = prec, groupName = grp,
                                 MSPeakLists = rmdVars$MSPeakLists, onlyAnnotated = TRUE)
        if (is.null(apl) || nrow(apl) == 0)
            return("<div align=\"center\">No annotation available.</div>")
        
        apl[, formula := subscriptFormulaHTML(formula)]
        apl[, neutral_loss := subscriptFormulaHTML(neutral_loss)]
        
        knitr::kable(apl, "html", escape = FALSE) %>%
            kableExtra::kable_styling(font_size = 11) %>%
            kableExtra::scroll_box(extra_css = "overflow: auto; height: 125px;")
    })
    
    ret <- data.table(group = match(grp, names(rmdVars$fGroups)),
                      formula = subscriptFormulaHTML(precursors),
                      spectrum = paste0(imgTags(allPlots[[grp]]$formulaSpecs), "<br>", fiTables),
                      scores = paste0(imgTags(allPlots[[grp]]$formulaScores), "<br>", infoTexts))

    return(ret)
}))

DT::datatable(formulasDT, options = list(scrollX = TRUE, scrollY = "600px", deferRender = TRUE,
                                         dom = "lrtp", pageLength = 25, autoWidth = TRUE,
                                         ordering = FALSE,
                                         columnDefs = list(list(visible = FALSE, targets = 0))),
              rownames = FALSE, escape = FALSE, elementId = "formulasTable")

```



```{r echo=FALSE, eval=length(compsClustFGroups) > 0}
rmdTexts <- vector("character", length = length(compsClustFGroups))

message("Generating compounds cluster layout... ", appendLF = FALSE)
# prog <- openProgBar(0, length(plotGroups))
compClustTempl <- readAllFile(system.file("templates", "comp-cluster.Rmd", package = "patRoon"))

cutcl <- cutClusters(rmdVars$compsCluster)
for (i in seq_along(compsClustFGroups))
{
    grp <- compsClustFGroups[i]
    grpi <- match(grp, names(rmdVars$fGroups))
    
    ct <- cutcl[[grp]]
    rmdTexts[i] <-
        paste0(glue::glue(compClustTempl,
                          grpi = grpi,
                          grp = grp,
                          dendro = allPlots[[grp]]$compClustDendro,
                          mcs = paste0(sprintf("![](%s)", allPlots[[grp]]$compClustMCS),
                                       collapse = "\n")),
               collapse = "\n")
}

rmdText <- paste0(rmdTexts, collapse = "\n")
message("Done!")
```

`r if (length(compsClustFGroups) > 0) rmdText`
