Details {data-orientation=rows}
===

<label for="view-select">View:</label>
<select name="view" id="view-select">
    <option value="plain">Plain</option>
    <option value="suspects">Suspects</option>
</select>

## { data-height=150 }

### Feature groups { data-width=200 }

```{r}
reactable::reactable(as.data.table(rmdVars$fGroups, qualities = "score", average = TRUE))
```

##
### Details component/parent/suspect { data-width=100 }

```{r}
plot(1:10)
```

## { data-height=200, .tabset }

### Feature group

```{r}
plot(1:10)
```

### Features

### Formulas

### Compounds

### Transformations




Details2 {data-orientation=rows}
===

<label for="view-select">View:</label>
<select name="view" id="view-select">
    <option value="plain">Plain</option>
    <option value="suspects">Suspects</option>
</select>

## { data-height=250 }

### Feature groups { data-width=250 }

```{r}
tab <- as.data.table(rmdVars$fGroups, qualities = "score", average = TRUE)
tab[, ret := round(ret, 2)]
tab[, mz := round(mz, 5)]

groupBy <- NULL
if (F && !is.null(rmdVars$components))
{
    tabc <- as.data.table(rmdVars$components)
    tab <- merge(tabc, tab)
    groupBy <- "name"
}
#reactable::reactable(tab, groupBy = groupBy)

ci <- copy(componentInfo(rmdVars$components))
setnames(ci, sub("^parent_", "", names(ci)))
ci[, c("SMILES", "InChI", "InChIKey", "size", "links") := NULL]
ci[, rt := round(rt, 2)]
ci[, c("mz", "neutralMass") := .(round(mz, 5), round(neutralMass, 5))]
reactable::reactable(ci, details = function(index)
{
    htmltools::div(style = "padding: 1rem",
                   reactable::reactable(tab[group %in% rmdVars$components[[index]]$group],
                                        pagination = FALSE, outlined = TRUE,
                                        theme = reactable::reactableTheme(backgroundColor = "grey")))
}, pagination = FALSE, onClick = "expand")

```

### Component EIC { data-width=100 }

<div>
<input type="radio" id="eic-choice" name="componentView" value="eic">
<label for="eic-choice">EIC</label>
<input type="radio" id="spec-choice" name="componentView" value="spec">
<label for="spec-choice">Spectrum</label>
</div>

<div>
<button type="button">EIC</button>
<button type="button">Spectrum</button>
</div>

```{r fig.show='hold'}
plot(1:10)
```

## { data-height=250, .tabset }

### Feature group

```{r}
plot(1:10)
```

### Features

### Formulas

### Compounds

### Transformations
