Details {data-orientation=rows}
===

<label for="view-select">View:</label>
<select name="view" id="view-select">
    <option value="plain">Plain</option>
    <option value="suspects">Suspects</option>
</select>

## { data-height=150 }

### Feature groups { data-width=200 }

```{r}
reactable::reactable(as.data.table(rmdVars$fGroups, qualities = "score", average = TRUE))
```

##
### Details component/parent/suspect { data-width=100 }

```{r}
plot(1:10)
```

## { data-height=200, .tabset }

### Feature group

```{r}
plot(1:10)
```

### Features

### Formulas

### Compounds

### Transformations


Details2 {data-orientation=columns}
===

<label for="view-select">View:</label>
<select name="view" id="view-select">
    <option value="plain">Plain</option>
    <option value="suspects">Suspects</option>
</select>

## { data-width=150, .tabset }

### components/parents/suspects

```{r}
reactable::reactable(as.data.table(rmdVars$fGroups, qualities = "score", average = TRUE))
```

### Heatmap

### Dendro

### plotGraph

## { data-width=250 }

### Feature groups

```{r}
reactable::reactable(as.data.table(rmdVars$fGroups, qualities = "score", average = TRUE))
```


## { data-width=200, .tabset }

### Feature group

```{r}
plot(1:10)
```

### Features

### Formulas

### Compounds

### Transformations
