Details {data-orientation=columns}
===

<script>
function updateView(sel)
{
    for (var o of document.getElementById("view-select").options)
        document.getElementById("detailsTab" + o.value).style.display = (o.value === sel) ? "" : "none"
}
</script>

<label for="view-select">View:</label>
<select name="view" id="view-select" onChange="updateView(this.value)">
    <option value="Plain">Plain</option>
    <option value="Suspects">Suspects</option>
    <option value="Components">Components</option>
    <option value="TPs">Transformation products</option>
</select>

## { data-width=200 }

### Feature groups { data-width=250 }

```{r echo=FALSE,eval=TRUE}
generator$genFeatTablePlain()
generator$genFeatTableSuspects()
generator$genFeatTableComponents()
generator$genFeatTableTPs()
```

## { data-width=150 }

### Selected parent { data-width=100 }

<script>
function updateParentImg(sel)
{
    let elements = document.getElementsByName("parentImgs");
    for (var i=0; i<elements.length; i++)
        document.getElementById(elements[i].value + "_view").style.display = (elements[i].value === sel) ? "" : "none";
}
</script>

<div>
<input type="radio" id="chrom-choice" name="parentImgs" value="chrom" onChange="updateParentImg(this.value);">
<label for="chrom-choice">Chromatogram</label>
<input type="radio" id="struct-choice" name="parentImgs" value="struct" onChange="updateParentImg(this.value);">
<label for="struct-choice">Structure</label>
<input type="radio" id="graph-choice" name="parentImgs" value="graph" onChange="updateParentImg(this.value);">
<label for="graph-choice">Graph</label>
</div>

<div>

<img id="chrom_view"></img>

<img id="struct_view"></img>

<div id="graph_view" style="height: 200px;">

```{r, echo = FALSE}
generator$genTPGraphs()
```

</div>

</div>

### Selected feature group

#### Feature group

```{r}
plot(1:10)
```

#### Features

#### Formulas

#### Compounds

#### Transformations
