Details {data-orientation=rows}
===

<script>
function updateView(sel)
{
    for (var o of document.getElementById("view-select").options)
        document.getElementById("detailsTab" + o.value).style.display = (o.value === sel) ? "" : "none"
}
</script>

<label for="view-select">View:</label>
<select name="view" id="view-select" onChange="updateView(this.value)">
    <option value="Plain">Plain</option>
    <option value="Suspects">Suspects</option>
    <option value="Components">Components</option>
    <option value="TPs">Transformation products</option>
</select>

## { data-height=250 }

### Feature groups { data-width=250 }

```{r echo=FALSE,eval=TRUE}
generator$genFeatTablePlain()
generator$genFeatTableSuspects()
generator$genFeatTableComponents()
generator$genFeatTableTPs()

```

### Parent { data-width=100 }

<div>
<input type="radio" id="eic-choice" name="componentView" value="eic">
<label for="eic-choice">EIC</label>
<input type="radio" id="spec-choice" name="componentView" value="spec">
<label for="spec-choice">Spectrum</label>
</div>

<div>
<button type="button">EIC</button>
<button type="button">Spectrum</button>
</div>

<div style="flex-wrap: nowrap;">

#### EIC

<img id="chrom_view"></img>

#### Structure

<img id="struct_view"></img>

#### Transformations

<div style="height: 200px;">

```{r, echo = FALSE}
generator$genTPGraphs()
```

</div>

</div>

## { data-height=250, .tabset }

### Feature group

```{r}
plot(1:10)
```

### Features

### Formulas

### Compounds

### Transformations
