---
title: "report"
author: "`r getPackageName(create = FALSE)`"
date: "`r if (!rmdVars$noDate) date() else '' `"
output:
    flexdashboard::flex_dashboard:
        vertical_layout: fill
        mathjax: null
        fig_mobile: false
        dev: png
---

```{r setup, include=FALSE}
# knitr::knit_hooks$set(optipng = knitr::hook_pngquant)
knitr::knit_hooks$set(pngquant = function(before, options, envir) suppressMessages(knitr::hook_pngquant(before, options, envir)))
knitr::opts_chunk$set(echo = FALSE, fig.keep = "all", fig.retina = 1, dpi = 72)
if (rmdVars$optimizePng)
    knitr::opts_chunk$set(pngquant = "")

# utility funcs for plotting
getPlotPath <- function(link)
{
    if (rmdVars$selfContained)
        ret <- "."
    else if (link)
        ret <- file.path("report_files", "plots")
    else
        ret <- file.path(rmdVars$outPath, "report_files", "plots")
    mkdirp(ret)
    return(ret)
}

imgTags <- function(img, style = "") 
{
    if (rmdVars$selfContained)
        ret <- sprintf("<img src=%s style='%s'></img>", img, style)
    else
    {
        # return(sprintf("<img src=file://%s></img>", img))
        ret <- sprintf("<img src='%s' style='%s'></img>", img, style)
    }
    return(ifelse(nzchar(img), ret, ""))
}

makeInfoBox <- function(txt)
{
    sprintf("<div style='max-width: 300px; max-height: 432px; border: 1px solid black; border-style: dotted; margin: 1px; padding: 1px; overflow: auto; white-space: nowrap; text-align: left;'>%s</div>", txt)
}

rGroupLenNonEmpty <- length(unique(analysisInfo(removeEmptyAnalyses(rmdVars$fGroups))$group))
rGroupLen <- length(replicateGroups(rmdVars$fGroups))
anyOverlap <- rGroupLen > 1 &&
    length(unique(rmdVars$fGroups, which = replicateGroups(rmdVars$fGroups), outer = TRUE)) < length(rmdVars$fGroups)
if (length(rmdVars$fGroups) > 0 && anyOverlap && rGroupLenNonEmpty > 1)
{
    doPlotChord <- "chord" %in% rmdVars$reportPlots && rGroupLenNonEmpty > 2
    doPlotVenn <- "venn" %in% rmdVars$reportPlots && rGroupLen < 6
    doPlotUpSet <- "upset" %in% rmdVars$reportPlots
} else
    doPlotChord <- doPlotVenn <- doPlotUpSet <- FALSE
doAnnotation <- !is.null(rmdVars$compounds) || !is.null(rmdVars$compsCluster) || !is.null(rmdVars$formulas) ||
    !is.null(rmdVars$components) || inherits(rmdVars$fGroups, "featureGroupsScreening")
doEICs <- length(rmdVars$fGroups) > 0 && "eics" %in% rmdVars$reportPlots
doIS <- nrow(internalStandards(rmdVars$fGroups)) > 0
isComponentsTP <- !is.null(rmdVars$components) && inherits(rmdVars$components, "componentsTPs")
rmdText <- NULL
```

<style>
pre, code {
    white-space: pre !important;
    overflow-x: auto !important;
    max-height: 275px;
    overflow-y: auto;
}
</style>

```{r echo=FALSE,eval=doEICs || doAnnotation}
message("Generating chromatograms...")

plotPathFull <- getPlotPath(FALSE)

prog <- openProgBar(0, length(rmdVars$fGroups))
allPlots <- sapply(seq_len(length(rmdVars$fGroups)), function(grpi)
{
    path <- file.path(plotPathFull, sprintf("chrom_%d.png", grpi))
    makeCachedPlot(path, "plotChroms",
                   list(rmdVars$fGroups[, grpi], rmdVars$EICRtWindow,
                        rmdVars$EICMzExpWindow, rmdVars$retMin,
                        rmdVars$EICTopMost, rmdVars$EICTopMostByRGroup, rmdVars$EICs, colourBy = "rGroup",
                        onlyPresent = rmdVars$EICOnlyPresent),
                   7, 4.5, bg = NA, cacheDB = rmdVars$cacheDB)
    setTxtProgressBar(prog, grpi)
    return(path)
})
close(prog)

if (rmdVars$optimizePng && length(allPlots) > 0)
    optimizePngPlots(allPlots)

chromPaths <- file.path(getPlotPath(TRUE), sprintf("chrom_%d.png", seq_len(length(rmdVars$fGroups))))
chromPathsFull <- file.path(plotPathFull, sprintf("chrom_%d.png", seq_len(length(rmdVars$fGroups))))
if (rmdVars$selfContained)
    chromPaths <- sapply(chromPaths, knitr::image_uri)

# stuff everything together: https://stackoverflow.com/a/21730473
rmdText <- sprintf("<script>var chromPaths = [ %s ];</script>",
                   paste0("'", chromPaths, "'", collapse = ", "))
```
`r if (!is.null(rmdText)) rmdText`

Summary {data-orientation=rows}
================

## { data-height=350 }

### EICs

```{r obj-plot, fig.width=10, fig.height=4}
par(mai = c(0.9, 0.8, 0.6, 0.1))
plotChroms(rmdVars$fGroups, rmdVars$EICRtWindow, rmdVars$EICMzExpWindow,
           rmdVars$retMin, 1, FALSE, rmdVars$EICs, TRUE, FALSE, colourBy = "fGroups", showLegend = FALSE,
           onlyPresent = rmdVars$EICOnlyPresent)
```


## { data-height=300 }

### Objects

```{r obj-show}
objToShow <- list(rmdVars$fGroups, rmdVars$MSPeakLists, rmdVars$formulas,
                  rmdVars$compounds, rmdVars$components)
objToShow <- objToShow[!sapply(objToShow, is.null)]
for (obji in seq_along(objToShow))
{
    show(objToShow[[obji]])
    cat("\n")
}
```


### Retention vs m/z
```{r fig.height=4}
par(mai = c(0.9, 0.8, 0.1, 0.1))
plot(rmdVars$fGroups, colourBy = "fGroups", showLegend = FALSE, retMin = rmdVars$retMin)
```


`r if (doPlotChord || doPlotVenn || doPlotUpSet) "## { data-height=425 } \n"`

`r if (doPlotChord) "### Chord diagram\n"`
```{r fig.height=5.5, eval=doPlotChord,out.height="400px"}
message("Creating chord diagram... ", appendLF = FALSE)
plotChord(rmdVars$fGroups, average = TRUE)
message("Done!")
```

`r if (doPlotVenn) "### Venn diagram\n"`
```{r fig.height=5.5, eval=doPlotVenn}
plotVenn(rmdVars$fGroups)
```

`r if (doPlotUpSet) "### UpSet diagram\n"`
```{r fig.height=5.5, eval=doPlotUpSet}
plotUpSet(rmdVars$fGroups)
```

`r if (doEICs) "EICs {data-navmenu=\"Features\"}\n===\n"`
```{r results='asis', eval=doEICs}
cat(sprintf("![%s](%s)", names(rmdVars$fGroups), chromPaths), sep = "\n")
```


```{r child="featinfo.Rmd", eval=doEICs}
```

`r if (doIS) "Internal standard assignments {data-navmenu=\"Features\"}\n===\n"`
```{r eval=doIS}
# UNDONE: set onlyPresent?

rmdText <- if (isFGSet(rmdVars$fGroups))
{
    paste0("## Plots {.tabset}\n\n", paste0(sapply(sets(rmdVars$fGroups), function(s)
    {
        glue::glue("### { s } {{ data-height=600 }}\n\n",
                   "{ ticks } {{r}}\nplotGraph(rmdVars$fGroups, set = \"{ s }\")\n{ ticks }\n\n",
                   ticks = "```", s = s)
    }), collapse = ""))
} else
    glue::glue("##\n\n### {{ data-height=600 }}\n\n",
               "{ ticks }{{r}}\nplotGraph(rmdVars$fGroups)\n{ ticks }\n\n", ticks = "```")
```

`r if (doIS) paste0(knitr::knit(text = rmdText), collapse = "\n")`


```{r child="components.Rmd", eval=!is.null(rmdVars$components) && length(rmdVars$components) > 0 && !isComponentsTP}
```


```{r child="annotation.Rmd", eval=doAnnotation }
```

```{r child="TPs.Rmd", eval=!is.null(rmdVars$components) && length(rmdVars$components) > 0 && isComponentsTP}

```{r child="details.Rmd"}
