---
title: "report"
author: "`r getPackageName(create = FALSE)`"
date: "`r if (!properties$noDate) date() else '' `"
output:
    flexdashboard::flex_dashboard:
        vertical_layout: fill
        mathjax: null
        fig_mobile: false
        dev: svg
        theme:
            version: 5
            bootswatch: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.keep = "all", fig.retina = 1, dpi = 72)
```

<style>
pre, code {
    white-space: pre !important;
    overflow-x: auto !important;
    height: auto;
    overflow-y: auto;
}

/* set positioning to allow scaling: https://stackoverflow.com/a/45144857 */
/* UNDONE: doesn't work...
.svglite {
    position: absolute;
    width: 100%;
    height: auto;
}*/

/* Image zooming, based on https://stackoverflow.com/a/57694495 */
.zoomDiv {
    opacity: 0;
    position: absolute;
    top: 50%;
    left: 50%;
    z-index: 2000;
    transform: translate(-50%, -50%);
    box-shadow: 0px 0px 50px #888888;
    max-height: 100%; 
    overflow: auto;
    cursor: zoom-out;
}

.zoomImg {
    width: 100%;
}

img:not(.zoomImg) {
    cursor: zoom-in;
    max-width: 100%;
    max-height: 100%;
}

</style>

<script type="text/javascript">
    // Image zooming, based on https://stackoverflow.com/a/57694495
    $(document).ready(function() {
        $('body').prepend("<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>");
        $('body').on('click', 'img:not(.zoomImg)', function() {
           $('.zoomImg').attr('src', $(this).attr('src'));
           $('.zoomDiv').css({opacity: '1', width: '70%'});
        });
        $('img.zoomImg').click(function() {
           $('.zoomDiv').css({opacity: '0', width: '0%'});
        });
    });
</script>

Summary
================

##

```{r, echo=FALSE}
rightColumnCards <- list(
    bslib::card(
        bslib::card_header("Objects"),
        bslib::card_body_fill(htmltools::pre(htmltools::code(objectsShow)))
    )
)

if (!is.null(plots$overview$chord) || !is.null(plots$overview$venn) || !is.null(plots$overview$UpSet))
{
    tabs <- list()
    makeTab <- function(title, plot) bslib::nav(title, bslib::card_body_fill(htmltools::HTML(generator$plotImg(plot))))
    if (!is.null(plots$overview$chord))
        tabs <- c(tabs, list(makeTab("Chord diagram", plots$overview$chord)))
    if (!is.null(plots$overview$venn))
        tabs <- c(tabs, list(makeTab("Venn diagram", plots$overview$venn)))
    if (!is.null(plots$overview$UpSet))
        tabs <- c(tabs, list(makeTab("UpSet diagram", plots$overview$UpSet)))
    rightColumnCards <- c(rightColumnCards,
                          list(style = "grid-template-rows: 2fr 3fr;",
                               do.call(bslib::navs_tab_card, c(list(title = "Feature distribution"), tabs))))
}

bslib::layout_column_wrap(
    width = 1/2,
    height = "100%",
    heights_equal = "row",
    style = "padding-bottom: 10px; padding-right: 10px; grid-template-columns: 2fr 1fr;",
    bslib::layout_column_wrap(
        width = 1,
        heights_equal = "row",
        bslib::card(
            bslib::card_header("Most intense feature of each group"),
            bslib::card_body_fill(htmltools::HTML(generator$plotImg(plots$overview$chroms)))
        ),
        bslib::card(
            bslib::card_header("Retention vs m/z"),
            bslib::card_body_fill(htmltools::HTML(generator$plotImg(plots$overview$retMZ)))
        )
    ),
    do.call(bslib::layout_column_wrap, c(list(
        width = 1,
        heights_equal = "row"
    ), rightColumnCards))
)
```

```{r child="details.Rmd"}
```
