# Advanced usage

## Feature parameter optimization

Many different parameters exist that may affect the output quality of feature finding and grouping. To avoid time consuming manual experimentation, functionality is provided to largely automate the optimization process. The methodology, which uses design of experiments (DoE), is based on the excellent [Isotopologue Parameter Optimization (IPO)](IPO) R package. The functionality of this package is directly integrated in patRoon. Some functionality was added or changed, the most important being support for other feature finding and grouping algorithms besides [XCMS] and basic optimization support for qualitative parameters. Nevertheless, the core optimization algorithms are largely untouched.

This section will introduce the most important concepts and functionalities. Please see the reference manual for more information (e.g. `` ?`feature-optimization` ``).

### Parameter sets

Before starting an optimization experiment we have to define _parameter sets_. These sets contain the parameters and (initial) numeric ranges that should be tested. A parameter set is defined as a regular `list`, and can be easily constructed with the `generateFeatureOptPSet()` and  `generateFGroupsOptPSet()` functions (for feature finding and feature grouping, respectively).

```{r fOptPSet,eval=FALSE}
pSet <- generateFeatureOptPSet("openms") # default test set for OpenMS
pSet <- generateFeatureOptPSet("openms", chromSNR = c(5, 10)) # add parameter
# of course manually making a list is also possible (e.g. if you don't want to test the default parameters)
pSet <- list(noiseThrInt = c(1000, 5000))
```

When a parameter set has been defined it should be used as input for the `optimizeFeatureFinding()` or `optimizeFeatureGrouping()` functions.

```{r fDoOpt,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo, "openms", pSet)
fgOpt <- optimizeFeatureGrouping(fList, "openms", pSet) # fList is an existing features object
```

Similar to `findFeatures()`, the first argument to `optimizeFeatureFinding()` should be the [analysis information](#anaInfo). Note that it is not uncommon to perform the optimization with only a subset of (representative) analyses (i.e. to reduce processing time).

```{r fDoOpt2,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo[1:2, ], "openms", pSet) # only use first two analyses
```

From the parameter set a design of experiment will be automatically created. Obviously, the more parameters are specified, the longer such an experiment will take. After an experiment has finished, the optimization algorithm will start a new experiment where numeric ranges for each parameter are increased or decreased in order to more accurately find optimum values. Hence, the numeric ranges specified in the parameter set are only _initial_ ranges, and will be changed in subsequent experiments. After each experiment iteration the results will be evaluated and a new experiment will be started as long as better results were obtained during the last experiment (although there is a hard limit defined by the `maxIterations` argument).

For some parameters it is recommended or even necessary to set hard limits on the numeric ranges that are allowed to be tested. For instance, setting a minimum feature intensity threshold is highly recommended to avoid excessive processing time and potentially suboptimal results due to excessive amounts of resulting features. Configuring absolute parameter ranges is done by setting the `paramRanges` argument.

```{r fOptPRanges,eval=FALSE}
# set minimum intensity threshold (but no max)
ftOpt <- optimizeFeatureFinding(anaInfo, "openms", pSet,
                                paramRanges = list(noiseThrInt = c(500, Inf)))
```

Depending on the used algorithm, several default absolute limits are imposed. These may be obtained with the `getDefFeaturesOptParamRanges()` and `getDefFGroupsOptParamRanges()` functions. 

The common operation is to optimize numeric parameters. However, parameters that are not numeric (i.e. _qualitative_) need a different approach. In this case you will need to define multiple parameter sets, where each set defines a different qualitative value.

```{r fOptQual,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo, "openms",
                                list(chromFWHM = c(4, 8), isotopeFilteringModel = "metabolites (5% RMS)"),
                                list(chromFWHM = c(4, 8), isotopeFilteringModel = "metabolites (2% RMS)"))
```

In the above example there are two parameter sets: both define the numeric `chromFWHM` parameter, whereas the qualitative `isotopeFilteringModel` parameter has a different value for each. Note that we had to specify the `chromFWHM` twice, this can be remediated by using the `templateParams` argument:

```{r fOptQualT,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo, "openms",
                                list(isotopeFilteringModel = "metabolites (5% RMS)"),
                                list(isotopeFilteringModel = "metabolites (2% RMS)"),
                                templateParams = list(chromFWHM = c(4, 8)))
```

As its name suggests, the `templateParams` argument serves as a template parameter set, and its values are essentially combined with each given parameter set. Note that current support for optimizing qualitative parameters is relatively basic and time consuming. This is because tests are essentially repeated for each parameter set (e.g. in the above example the `chromFWHM` parameter is optimized twice, each time with a different value for `isotopeFilteringModel`).

### Processing optmization results

The results of an optimization process are stored in objects from the S4 `optimizationResult` class. Several methods are defined to inspect and visualize these results.

The `optimizedParameters()` function is used to inspect the best parameter settings. Similarly, the `optimizedObject()` function can be used to obtain the object created with these settings (i.e. a `features` or `featureGroups` object).

```{r fOptProc,eval=FALSE}
optimizedParameters(ftOpt) # best settings for whole experiment
optimizedParameters(ftOpt, 1) # best settings for first parameter set
optimizedParameters(ftOpt, 1, 1) # best settings for first parameter set and experiment iteration

optimizedObject(ftOpt) # features object with best settings for whole experiment
optimizedObject(ftOpt, 1) # features object with best settings for first parameter set
```

The `plot()` function can be used to visualize optimization results. This function will plot graphs for results of all tested parameter pairs. The graphs can be contour, image or perspective plots (as specified by the `type` argument).

```{r fOptPlot,eval=FALSE}
plot(ftOpt, paramSet = 1, DoEIteration = 1) # contour plots for first param set/experiment
plot(ftOpt, paramSet = 1, DoEIteration = 1, type = "persp") # pretty perspective plots
```

Please refer to the reference manual for more methods to inspect optimization results (e.g. `?optimizationResult`).

## Algorithm consensus {#consensus}

- fGroup comparison
- consensus selection (overlap/unique)

## Compound clustering {#compclust}

- Principle
- Cluster selection, MCS plotting
- methods

## Caching

## Parallelization
