# Installation

`patRoon` depends on various other software tools to perform the non-target analysis workflow steps and to implement various other functionalities. Most of these dependencies are automatically installed when you install the `patRoon` R package, however, some need to be manually installed and/or configured. 

> **_NOTE_**  It is highly recommended to perform installation steps in a 'clean' `R` session to avoid errors when installing or upgrading packages. As such it is recommended to close all open (R Studio) sessions and open a plain R console to perform the installation. 
 
## Automatic installation (Windows only)

An installation script is provided that automatically installs and configures all dependencies and finally installs `patRoon` itself. At this moment, this script only works with Microsoft Windows. You don't have to install anything else to use it, simply open `R` and execute these commands:

```{r eval=FALSE}
source("https://raw.githubusercontent.com/rickhelmus/patRoon/master/install_patRoon.R")
installPatRoon()
```

A simple text based wizard will start and asks you what to install and how to do it. You can re-run this installer at any time, for instance, if something went wrong or you want to install additional dependencies.

## Manual installation

Alternatively, the manual installation is for users who don't use Windows, prefer to do a manual installation or simply want to know what happens behind the scenes. The manual installation consists of three phases:

1. Installing some prerequisite `R` packages
2. Install and configure (non-`R`) dependencies
3. Install `patRoon`

### R prerequisites

When installing `patRoon` Windows users have the option to install from a customized ([miniCRAN]) repository (`patRoonDeps`). This repository provides a central repository for `patRoon` and all its R packages. An advantage is that installation will be faster and you will not need to install [Rtools]. Note that you will need to have the latest `R` version installed in order to use this repository.

When you decide to use the `patRoonDeps` repository you can simply _skip_ this step. **Otherwise** (i.e. you will use regular repositories instead), execute the following:

```{r eval=FALSE}
install.packages(c("BiocManager", "remotes"))
BiocManager::install("CAMERA")

# only needed for Bruker DataAnalysis integration
install.packages("RDCOMClient", repos = "http://www.omegahat.net/R")

# only when using the R interface (not recommended by default)
remotes::install_github("c-ruttkies/MetFragR/metfRag")
```

Note that the latter two commands concern installation of _optional_ packages. If you are unsure, you probably don't need them and can always install them later.

### Other dependencies

Depending on which functionality is used, the following external dependencies may be installed:

Software                         | Remarks
-------------------------------- | -----------------------------------------------------
[Java JDK][JavaJDK]              | **Mandatory** for e.g. plotting structures and using MetFrag.
[Rtools]                         | Necessary on Window and when `patRoon` is _not_ installed from `patRoonDeps`.
[ProteoWizard]                   | Needed for automatic data-pretreatment (e.g. data file conversion and centroiding, Bruker users may use DataAnalysis integration instead).
[OpenMS]                         | Recommended. Used for e.g. finding and grouping features.
[MetFrag CL][MetFragCL]          | Recommended. Used for annotation with MetFrag.
[MetFrag CompTox DB][CompTox-dl] | Database files necessary for usage of the [CompTox] database with MetFrag. Note that currently only scorings from the `SelectMetaData` and `SelectMetaDataPlus` files are supported and a recent version (>=2.4.5) of MetFrag is required.
[SIRIUS]                         | For formula and/or compound annotation.
[pngquant]                       | Used to reduce size of HTML reports, definitely optional.

After installation you may need to configure the file path to ProteoWizard, OpenMS, SIRIUS, MetFrag, the MetFrag CompTox DB and/or pngquant (normally ProteoWizard and OpenMS should be automatically found). To configure their file locations you should set some global package options with the `option()` R function, for instance:

```{r, eval=FALSE}
options(patRoon.path.pwiz = "C:/ProteoWizard") # location of ProteoWizard installation folder
options(patRoon.path.SIRIUS = "C:/sirius-win64-3.5.1") # location where SIRIUS was extracted
options(patRoon.path.OpenMS = "/usr/local/bin") # directory with the OpenMS binaries
options(patRoon.path.pngquant = "~/pngquant") # directory containing pngquant binary
options(patRoon.path.MetFragCL = "~/MetFrag2.4.5-CL.jar") # full location to the jar file
options(patRoon.path.MetFragCompTox = "C:/DSSTox_01May18_Full_SelectMetaDataPlus.csv") # full location to desired CompTox CSV file
```

These commands have to be executed everytime you start a new R session (e.g. as part of your script). However, it is probably easier to add them to your `~/.Rprofile` file so that they are executed automatically when you start R. If you don't have this file yet you can simply create it yourself (for more information see e.g. [this SO answer](https://stackoverflow.com/a/46819910)).

### patRoon installation

Finally, it is time to install `patRoon` itself. As mentioned before, Windows users (who have the latest `R` version) can install `patRoon` and all its package dependencies from the `patRoonDeps` repository:

```{r eval=FALSE}
# install from patRoonDeps (only Windows with latest R version)
install.packages("patRoon", repos = "https://rickhelmus.github.io/patRoonDeps/", type = "binary")

# optional, data for tutorial
install.packages("patRoonData", repos = "https://rickhelmus.github.io/patRoonDeps/", type = "binary")
```

**Otherwise**, installation occurs directly from GitHub:

```{r eval=FALSE}
remotes::install_github("rickhelmus/patRoon")

# optional, data for tutorial
remotes::install_github("rickhelmus/patRoonData") # example data used by tutorial
```

Afterwards, you can run the `verifyDependencies()` function to see if `patRoon` can find all its dependencies (you may need to restart R beforehand)

```{r eval=FALSE}
patRoon::verifyDependencies()
```
