# Advanced usage

## Adducts {#adducts}

When generating formulae and compound annotations and some other functionalities it is required to specify the adduct species. Behind the scenes, different algorithms typically use different formats. For instance, in order to specify a protonated species...

* `GenForm` either accepts `"M+H"` and `"+H"`
* `MetFrag` either accepts the numeric code `1` or `"[M+H]+"`
* `SIRIUS` accepts `"[M+H]+"`

In addition, most algorithms only accept a limited set of possible adducts, which do not necessarily all overlap with each other. The `GenFormAdducts()` and `MetFragAdducts()` functions list the possible adducts for `GenForm` and `MetFrag`, respectively.

In order to simplify the situation `patRoon` internally uses its own format and converts it automatically to the algorithm specific format when necessary. Furthermore, during conversion it checks if the specified adduct format is actually allowed by the algorithm. Adducts in `patRoon` are stored in the `adduct` S4 class. Objects from this class specify which elements are added and/or subtracted, the final charge and the number of molecules present in the adduct (e.g. _2_ for a dimer).

```{r adduct1,eval=FALSE}
adduct(add = "H") # [M+H]+
adduct(sub = "H", charge = -1) # [M-H]-
adduct(add = "K", sub = "H2", charge = -1) # [M+K-H2]-
adduct(add = "H3", charge = 3) # [M+H3]3+
adduct(add = "H", molMult = 2) # [2M+H]+
```

A more easy way to generate adduct objects is by using the `as.adduct()` function:

```{r adduct2,eval=FALSE}
as.adduct("[M+H]+")
as.adduct("[M+H2]2+")
as.adduct("[2M+H]+")
as.adduct("[M-H]-")
as.adduct("+H", format = "genform")
as.adduct(1, isPositive = TRUE, format = "metfrag")
```

In fact, the `adduct` argument to workflow functions such as `generateFormulas()` and `generateCompounds()` is automatically converted to an `adduct` class with the `as.adduct()` function if necessary:

```{r adduct3,eval=FALSE}
formulas <- generateFormulas(..., adduct = adduct(sub = "H", charge = -1))
formulas <- generateFormulas(..., adduct = "[M-H]-") # same as above
```

More details can be found in the reference manual (`?adduct` and ``?`adduct-utils` ``).

## Feature parameter optimization {#fOpt}

Many different parameters exist that may affect the output quality of feature finding and grouping. To avoid time consuming manual experimentation, functionality is provided to largely automate the optimization process. The methodology, which uses design of experiments (DoE), is based on the excellent [Isotopologue Parameter Optimization (IPO)](IPO) `R` package. The functionality of this package is directly integrated in patRoon. Some functionality was added or changed, the most important being support for other feature finding and grouping algorithms besides [XCMS] and basic optimization support for qualitative parameters. Nevertheless, the core optimization algorithms are largely untouched.

This section will introduce the most important concepts and functionality. Please see the reference manual for more information (e.g. `` ?`feature-optimization` ``).

> **_NOTE_** The [SIRIUS] and [SAFD] algorithms are currently not (yet) supported.

### Parameter sets

Before starting an optimization experiment we have to define _parameter sets_. These sets contain the parameters and (initial) numeric ranges that should be tested. A parameter set is defined as a regular `list`, and can be easily constructed with the `generateFeatureOptPSet()` and  `generateFGroupsOptPSet()` functions (for feature finding and feature grouping, respectively).

```{r fOptPSet,eval=FALSE}
pSet <- generateFeatureOptPSet("openms") # default test set for OpenMS
pSet <- generateFeatureOptPSet("openms", chromSNR = c(5, 10)) # add parameter
# of course manually making a list is also possible (e.g. if you don't want to test the default parameters)
pSet <- list(noiseThrInt = c(1000, 5000))
```

When optimizing with [XCMS] or [KPIC2] a few things have to be considered. First of all, when using the XCMS3 interface (i.e. `algorithm="xcms3"`) the underlying method that should be used for finding and grouping features and retention alignment should be set. In case these are not set default methods will be used.

```{r fOptPSetXCMS1,eval=FALSE}
pSet <- list(method = "centWave", ppm = c(2, 8))
pSet <- list(ppm = c(2, 8)) # same: centWave is default

# get defaults, but for different grouping/alignment methods
pSetFG <- generateFGroupsOptPSet("xcms3", groupMethod = "nearest", retAlignMethod = "peakgroups")
```

In addition, when optimizing feature grouping (both `XCMS` interfaces and `KPIC2`) we need to set the grouping and retention alignment parameters in two different nested lists: these are `groupArgs`/`retcorArgs` (`algorithm="xcms"`), `groupParams`/`retAlignParams` (`algorithm="xcms3"`) or `groupArgs`/`alignArgs` (`algorithm="kpic2"`).

```{r fOptPSetXCMS2,eval=FALSE}
pSetFG <- list(groupParams = list(bw = c(20, 30))) # xcms3
pSetFG <- list(retcorArgs = list(gapInit = c(0, 7))) # xcms
pSetFG <- list(groupArgs = list(mz_weight = c(0.3, 0.9))) # kpic2
```

When a parameter set has been defined it should be used as input for the `optimizeFeatureFinding()` or `optimizeFeatureGrouping()` functions.

```{r fDoOpt,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo, "openms", pSet)
fgOpt <- optimizeFeatureGrouping(fList, "openms", pSetFG) # fList is an existing features object
```

Similar to `findFeatures()`, the first argument to `optimizeFeatureFinding()` should be the [analysis information](#anaInfo). Note that it is not uncommon to perform the optimization with only a subset of (representative) analyses (i.e. to reduce processing time).

```{r fDoOpt2,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo[1:2, ], "openms", pSet) # only use first two analyses
```

From the parameter set a design of experiment will be automatically created. Obviously, the more parameters are specified, the longer such an experiment will take. After an experiment has finished, the optimization algorithm will start a new experiment where numeric ranges for each parameter are increased or decreased in order to more accurately find optimum values. Hence, the numeric ranges specified in the parameter set are only _initial_ ranges, and will be changed in subsequent experiments. After each experiment iteration the results will be evaluated and a new experiment will be started as long as better results were obtained during the last experiment (although there is a hard limit defined by the `maxIterations` argument).

For some parameters it is recommended or even necessary to set hard limits on the numeric ranges that are allowed to be tested. For instance, setting a minimum feature intensity threshold is highly recommended to avoid excessive processing time and potentially suboptimal results due to excessive amounts of resulting features. Configuring absolute parameter ranges is done by setting the `paramRanges` argument.

```{r fOptPRanges,eval=FALSE}
# set minimum intensity threshold (but no max)
ftOpt <- optimizeFeatureFinding(anaInfo, "openms",
                                list(noiseThrInt = c(1000, 5000), # initial testing range
                                paramRanges = list(noiseThrInt = c(500, Inf))) # never test below 500
```

Depending on the used algorithm, several default absolute limits are imposed. These may be obtained with the `getDefFeaturesOptParamRanges()` and `getDefFGroupsOptParamRanges()` functions. 

The common operation is to optimize numeric parameters. However, parameters that are not numeric (i.e. _qualitative_) need a different approach. In this case you will need to define multiple parameter sets, where each set defines a different qualitative value.

```{r fOptQual,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo, "openms",
                                list(chromFWHM = c(4, 8), isotopeFilteringModel = "metabolites (5% RMS)"),
                                list(chromFWHM = c(4, 8), isotopeFilteringModel = "metabolites (2% RMS)"))
```

In the above example there are two parameter sets: both define the numeric `chromFWHM` parameter, whereas the qualitative `isotopeFilteringModel` parameter has a different value for each. Note that we had to specify the `chromFWHM` twice, this can be remediated by using the `templateParams` argument:

```{r fOptQualT,eval=FALSE}
ftOpt <- optimizeFeatureFinding(anaInfo, "openms",
                                list(isotopeFilteringModel = "metabolites (5% RMS)"),
                                list(isotopeFilteringModel = "metabolites (2% RMS)"),
                                templateParams = list(chromFWHM = c(4, 8)))
```

As its name suggests, the `templateParams` argument serves as a template parameter set, and its values are essentially combined with each given parameter set. Note that current support for optimizing qualitative parameters is relatively basic and time consuming. This is because tests are essentially repeated for each parameter set (e.g. in the above example the `chromFWHM` parameter is optimized twice, each time with a different value for `isotopeFilteringModel`).

### Processing optmization results

The results of an optimization process are stored in objects from the S4 `optimizationResult` class. Several methods are defined to inspect and visualize these results.

The `optimizedParameters()` function is used to inspect the best parameter settings. Similarly, the `optimizedObject()` function can be used to obtain the object that was created with these settings (i.e. a `features` or `featureGroups` object).

```{r fOptProc1,eval=doOpt}
optimizedParameters(ftOpt) # best settings for whole experiment
optimizedObject(ftOpt) # features object with best settings for whole experiment
```

Results can also be obtained for specific parameter sets/iterations.

```{r fOptProc,eval=doOpt}
optimizedParameters(ftOpt, 1) # best settings for first parameter set
optimizedParameters(ftOpt, 1, 1) # best settings for first parameter set and experiment iteration
optimizedObject(ftOpt, 1) # features object with best settings for first parameter set
```

The `plot()` function can be used to visualize optimization results. This function will plot graphs for results of all tested parameter pairs. The graphs can be contour, image or perspective plots (as specified by the `type` argument).

```{r fOptPlot,eval=doOpt,out.width="70%"}
plot(ftOpt, paramSet = 1, DoEIteration = 1) # contour plots for first param set/experiment
plot(ftOpt, paramSet = 1, DoEIteration = 1, type = "persp") # pretty perspective plots
```

Please refer to the reference manual for more methods to inspect optimization results (e.g. `?optimizationResult`).

## Exporting and converting feature data

The feature group data obtained during the workflow can be exported to various formats with the `export()` generic function. There are currently three formats supported: `"brukerpa"` (Bruker ProfileAnalysis), `"brukertasq"` (Bruker TASQ) and `"mzmine"` (mzMine). The former exports a 'bucket table' which can be loaded in ProfileAnalysis, the second and third export a target list that can be processed with TASQ and mzMine, respectively.

The `getXCMSSet()` function converts a `features` or `featureGroups` object to an `xcmsSet` object which can be used for further processing with [xcms]. Similarly, the `getXCMSnExp()` function can be used for conversion to an XCMS3 style `XCMSnExp` object, and the `getPICSet()` function can be used to convert `features` to [KPIC2] data.

Some examples for these functions are shown below.

```{r fExpConv,eval=FALSE}
export(fGroups, "brukertasq", out = "my_targets.csv")

# convert features to xcmsSet.
# NOTE: loadRawData should only be FALSE when the analysis data files cannot be
# loaded by the algorithm (e.g. when features were obtained with DataAnalysis and data was not exported to mz(X)ML)
xset <- getXCMSSet(fList, loadRawData = TRUE)
xsetg <- getXCMSSet(fGroups, loadRawData = TRUE) # get grouped xcmsSet

# using the new XCMS3 interface
xdata <- getXCMSnExp(fList)
xdata <- getXCMSnExp(fGroups)

# KPIC2 conversion. Like XCMS it optionally loads the raw data.
picSet <- getPICSet(fList, loadRawData = TRUE)
```

## Algorithm consensus {#consensus}

With `patRoon` you have the option to choose between several algorithms for most workflow steps. Each algorithm is typically characterized by its efficiency, robustness, and may be optimized towards certain data properties. Comparing their output is therefore advantegeuous in order to design an optimum workflow. The `consensus()` generic function will compare different results from different algorithms and returns a _consensus_, which may be based on minimal overlap, uniqueness or simply a combination of all results from involved objects. The output from the `consensus()` function is of similar type as the input types and is therefore compatible to any 'regular' further data processing operations (e.g. input for other workflow steps or plotting). Note that a consensus can also be made from objects generated by the same algorithm, for instance, to compare or combine results obtained with different parameters (e.g. different databases used for compound annotation).

The `consensus()` generic is defined for most workflow objects. Some of its common function arguments are listed below.

Argument      | Classes                                            | Remarks
------------- | -------------------------------------------------- | -------------------------------------------------------------
`obj`, `...`  | All                                                | Two or more objects (of the same type) that should be compared to generate the consensus.
`compThreshold`, `relAbundance`, `absAbundance`, `formThreshold`   | `compounds`, `formulas`, `featureGroupsComparison` | The minimum overlap (relative/absolute) for a result (feature, candidate) to be kept.
`uniqueFrom`  | `compounds`, `formulas`, `featureGroupsComparison` | Only keep _unique_ results from specified objects.
`uniqueOuter` | `compounds`, `formulas`, `featureGroupsComparison` | Should be combined with `uniqueFrom`. If `TRUE` then only results are kept which are _also_ unique between the objects specified with `uniqueFrom`.

Note that current support for generating a consensus between `components` objects is very simplistic; here results are not compared, but the consensus simply consists a combination of all the components from each object.

Generating a consensus for feature groups involves first generating a `featureGroupsComparison` object. This step is necessary since (small) deviations between retention times and/or mass values reported by different feature finding/grouping algorithms complicates a direct comparison. The comparison objects are made by the `comparison()` function, and its results can be visualized by the plotting functions discussed [in the previous chapter](#visComp).

Some examples are shown below

```{r consensus,eval=FALSE}
compoundsCons <- consensus(compoundsMF, compoundsSIR) # combine MetFrag/SIRIUS results
compoundsCons <- consensus(compoundsMF, compoundsSIR,
                           compThreshold = 1) # only keep results that overlap

fGroupComp <- comparison(fGroupsXCMS, fGroupsOpenMS, fGroupsEnviPick,
                         groupAlgo = "openms")
plotVenn(fGroupComp) # visualize overlap/uniqueness
fGroupsCons <- consensus(fGroupComp,
                         uniqueFrom = 1:2) # only keep results unique in OpenMS+XCMS
fGroupsCons <- consensus(fGroupComp,
                         uniqueFrom = 1:2,
                         uniqueOuter = TRUE) # as above, but also exclude any overlap between OpenMS/XCMS
```

## Compound clustering {#compclust}

When large databases such as [PubChem] or [ChemSpider] are used for compound annotation, it is common to find _many_ candidate structures for even a single feature. While choosing the right scoring settings can significantly improve their ranking, it is still very much possible that many candidates of potential interest remain. In this situation it might help to perform _compound clustering_. During this process, all candidates for a feature are clustered hierarchically on basis of similar chemical structure. From the resulting cluster the _maximum common substructure_ (MCS) can be derived, which represents the largest possible substructure that 'fit' in all candidates. By visual inspection of the MCS it may be possible to identify likely common structural properties of a feature.

In order to perform compound clustering the `makeHCluster()` generic function should be used. This function heavily relies on chemical fingerprinting functionality provided by [rcdk].

```{r cClust,eval=FALSE}
compounds <- generateCompounds(...) # get our compounds
compsClust <- makeHCluster(compounds)
```

This function accepts several arguments to fine tune the clustering process:

* `method`: the clustering method (e.g. `"complete"` (default), `"ward.D2"`), see `?hclust` for options
* `fpType`: finger printing type (`"extended"` by default), see `?get.fingerprint`
* `fpSimMethod`: similarity method for generating the distance method (`"tanimoto"` by default), see `?fp.sim.matrix`

For all arguments see the reference manual (`?makeHClust`).

The resulting object is of type `compoundsCluster`. Several methods are defined to modify and inspect these results:

```{r cClustObj,eval=FALSE}
# plot MCS of first cluster from candidates of M109_R116_61
plotStructure(compsClust, groupName = "M109_R116_61", 1)

# plot dendrogram
plot(compsClust, groupName = "M109_R116_61")

# re-assign clusters for a feature group
compsClust <- treeCut(compsClust, k = 5, groupName = "M109_R116_61")
# ditto, but automatic cluster determination
compsClust <- treeCutDynamic(compsClust, minModuleSize = 3, groupName = "M109_R116_61")
```

For a complete overview see the reference manual (`?compoundsCluster`).

## Basic quantitative and regression analysis

While `patRoon` is currently primarily focused on qualitative analyses, some _basic_ quantitative analysis can also be performed, for instance, to estimate concentrations of features. In fact, other types of data that may be useful for regression analysis can be set such as sample dilution factor or sampling time. The latter may, for instance, be used to isolate features with increasing or decreasing intensity. Regardless of what kind of regression analysis is performed, here we simply refer the values to be calculated as _concentrations_. In order to use this functionality, an extra column (`conc`) should be added to the [analysis information](#anaInfo), for instance:

```{r quantAnaInfo,eval=FALSE}
# obtain analysis information as usual, but add some concentrations.
# The blanks are set to NA, whereas the standards are set to increasing levels.
anaInfo <- generateAnalysisInfo(paths = patRoonData::exampleDataPath(),
                                groups = c(rep("solvent", 3), rep("standard", 3)),
                                blanks = "solvent",
                                concs = c(NA, NA, NA, 1, 2, 3))
```

For analyses with known concentrations (e.g. standards) the `conc` column should be set; for all others the value should be set to `NA`.

The `as.data.table()` function (or `as.data.frame()`) can then be used to calculate regression data and estimate concentrations:

```{r quantDT,eval=FALSE}
# use areas for quantitation and make sure that feature data is reported
# (otherwise no concentrations are calculated)
# (only relevant columns are shown for clarity)
as.data.table(fGroups, areas = TRUE, features = TRUE, regression = TRUE)
```

```{r quantDTDo,echo=FALSE,eval=runData}
as.data.table(fGroupsConc, areas = TRUE, features = TRUE, regression = TRUE)[, c("group", "conc", "RSQ", "intercept", "slope", "conc_reg")]
```

Calculated concentrations are stored in the `conc_reg` column, alongside while other regression data (i.e. `RSQ`, `intercept`, `slope` columns). To perform basic trend analysis the `RSQ` (i.e. R squared) can be used:

```{r quantRSQ,eval=FALSE}
fGroupsTab <- as.data.table(fGroups, areas = TRUE, features = FALSE, regression = TRUE)
# subset fGroups with reasonable correlation
increasingFGroups <- fGroups[, fGroupsTab[RSQ >= 0.8, group]]
```

## Fold changes {#FCCalc}

A specific statistical way to prioritize feature data is by Fold changes (FC). This is a relative simple method to quickly identify (significant) changes between two sample groups. A typical use case is to compare the feature intensities before and after an experiment.

To perform FC calculations we first need to specify its parameters. This is best achieved with the `getFCParams()` function:

```{r FCParams,eval=TRUE}
getFCParams(c("before", "after"))
```

In this example we generate a `list` with parameters in order to make a comparison between two replicate groups: `before` and `after`. Several advanced parameters are available to tweak the calculation process. These are explained in the reference manual (`?featureGroups`).

The `as.data.table` function for feature groups is used to perform the FC calculations.

```{r include=FALSE,eval=runData}
fGroupsO <- fGroups
fGroups <- filter(fGroupsUF, absMinIntensity = 1E4)
```

```{r getFC,eval=runData}
myFCParams <- getFCParams(c("solvent-pos", "standard-pos")) # compare solvent/standard
as.data.table(fGroups, FCParams = myFCParams)[, c("group", "FC", "FC_log", "PV", "PV_log", "classification")]
```

The `classification` column allows you to easily identify if and how a feature changes between the two sample groups. This can also be used to prioritize feature groups:

```{r eval=FALSE}
tab <- as.data.table(fGroups, FCParams = myFCParams)
# only keep feature groups that significantly increase or decrease
fGroupsChanged <- fGroups[, tab[classification %in% c("increase", "decrease")]$group]
```

The `plotVolcano` function can be used to visually the FC data:

```{r plotVolcano,eval=runData}
plotVolcano(fGroups, myFCParams)
```

```{r include=FALSE,eval=runData}
fGroups <- fGroupsO
```

## Caching

In `patRoon` lengthy processing operations such as finding features and generating annotation data is _cached_. This means that when you run such a calculation again (without changing any parameters), the data is simply loaded from the cache data instead of re-generating it. This in turn is very useful, for instance, if you have closed your R session and want to continue with data processing at a later stage.

The cache data is stored in a [sqlite] database file. This file is stored by default under the name `cache.sqlite` in the current working directory (for this reason it is very important to always restore your working directory!). However, the name and location can be changed by setting a global package option:

```{r cacheFile,eval=FALSE}
options(patRoon.cache.fileName = "~/myCacheFile.sqlite")
```

For instance, this might be useful if you want to use a shared cache file between projects.

After a while you may see that your cache file can get quite large. This is especially true when testing different parameters to optimize your workflow. Furthermore, you may want to clear the cache after you have updated `patRoon` and want to make sure that the latest code is used to generate the data. At any point you can simply remove the cache file. A more fine tuned approach which doesn't wipe all your cached data is by using the `clearCache()` function. With this function you can selectively remove parts of the cache file. The function has two arguments: `what`, which specifies what should be removed, and `file` which specifies the path to the cache file. The latter only needs to be specified if you want to manage a different cache file.

In order to figure what is in the cache you can run `clearCache()` without any arguments:

```{r clearCache}
clearCache()
```

Using this output you can re-run the function again, for instance:

```{r clearCache2,eval=FALSE}
clearCache("featuresOpenMS")
clearCache(c("featureGroupsOpenMS", "formulasGenForm")) # clear multiple
clearCache("OpenMS") # clear all with OpenMS in name (ie partial matched)
clearCache("all") # same as simply removing the file
```

## Parallelization

Some steps in the non-target screening workflow are inherently computationally intensive. To reduce computational times `patRoon` is able to perform _parallelization_ for most of the important functionality. This is especially useful if you have a modern system with multiple CPU cores and sufficient RAM.

For various technical reasons several parallelization techniques are used, these can be categorized as _parallelization of `R` functions_ and _multiprocessing_. The next sections describe both parallelization approaches in order to let you optimize the workflow.

### Parellization of R functions

Several functions of `patRoon` support parallelization.

Function                 | Purpose                             | Remarks
------------------------ | ----------------------------------- | ------------------------
`findFeatures`           | Obtain feature data                 | Only `envipick` and `kpic2` algorithms
`generateComponents`     | Generate components                 | Only `cliquems` algorithm.
`optimizeFeatureFinding`, `optimizeFeatureGrouping` | Optimize feature finding/grouping parameters | Discussed [here](#fOpt).
`calculatePeakQualities` | Calculate feature (group) qualities | Discussed [here].

The parallelization is achieved with the [future] and [future.apply] `R` packages. To enable parallelization of these functions the `parallel` argument must be set to `TRUE` and the future framework must be properly configured in advance. For example:

```{r eval=FALSE}
# setup three workers to run in parallel
future::plan("multisession", workers = 3)

# find features with enviPick in parallel
fList <- findFeatures(anaInfo, "envipick", parallel = TRUE)
```

It is important to properly configure the right future plan. Please see the documentation of the [future] package for more details.

### Multiprocessing

`patRoon` relies on several external (command-line) tools to generate workflow data. These commands may be executed in _parallel_ to reduce computational times ('multiprocessing'). The table below outlines the tools that are executed in parallel.

Tool                  | Used by                                      | Notes
--------------------- | -------------------------------------------- | ---------------------------------
`msConvert`           | `convertMSFiles(algorithm="pwiz", ...)`      |
`FileConverter`       | `convertMSFiles(algorithm="openms", ...)`    |
`FeatureFinderMetabo` | `findFeatures(algorithm="openms", ...)`      |
`julia`               | `findFeatures(algorithm="safd", ...)`        |
`SIRIUS`              | `findFeatures(algorithm="sirius", ...)`      |
`MetaboliteAdductDecharger` | `generateComponents(algorithm="openms", ...)` |
`GenForm`             | `generateFormulas(agorithm="genform", ...)`  |
`SIRIUS`              | `generateFormulas(agorithm="sirius", ...)`, `generateCompounds(agorithm="sirius", ...)`  | Only if `splitBatches=TRUE`
`MetFrag`             | `generateCompounds(agorithm="metfrag", ...)` |
`pngquant`            | `reportHTML(...)`                            | Only if `optimizePng=TRUE`
`BioTransformer`      | `generateTPs(algorithm = "biotransformer")`  |

Multiprocessing is either performed by executing processes in the background with the [processx] `R` package (_classic interface_) or by futures, which were introduced in the previous section. An overview of the characteristics of both parallelization techniques is shown below.

`classic`                                           |  `future`
--------------------------------------------------- | -------------------------------------------------------
requires little or no configuration                 | configuration needed to setup
works with all tools                                | doesn't work with `pngquant` and slower with `GenForm`
only supports parallelization on the local computer | allows both local and cluster computing

Which method is used is controlled by the `patRoon.MP.method` package option. Note that `reportHTML()` will always use the classic method for `pngquant`.

#### Classic multiprocessing interface

The classic interface is the 'original' method implemented in `patRoon`, and is therefore well tested and optimized. It is easier to setup, works well with all tools, and is therefore the default method. It is enabled as follows:

```{r eval=FALSE}
options(patRoon.MP.method = "classic")
```

The number of parallel processes is configured through the `patRoon.MP.maxProcs` option. By default it is set to the number of available CPU cores, which results usually in the best performance. However, you may want to lower this, for instance, to keep your computer more responsive while processing or limit the RAM used by the data processing workflow.

```{r eval=FALSE}
options(patRoon.MP.maxProcs = 2) # do not execute more than two tools in parallel. 
```

This will change the parallelization for the complete workflow. However, it may be desirable to change this for only a part the workflow. This is easily achieved with the `withOpt()` function.

```{r eval=FALSE}
# do not execute more than two tools in parallel.
options(patRoon.MP.maxProcs = 2)

# ... but execute up to four GenForm processes
withOpt(MP.maxProcs = 4, {
    formulas <- generateFormulas(fGroups, "genform", ...)
})
```

The `withOpt` function will temporarily change the given option(s) while executing a given code block and restore it afterwards (it is very similar to the `with_options()` function from the [withr] `R` package). Furthermore, notice how `withOpt()` does not require you to prefix the option names with `patRoon.`. 

#### Multiprocessing with futures

The primary goal of the "future" method is to allow parallel processing on one or more external computers. Since it uses the [future] `R` package, many approaches are supported, such as local parallelization (similar to the `classic` method), cluster computing via multiple networked computers and more advanced HPC approaches such as `slurm` via the [future.batchtools] `R` package. This parallelization method can be activated as follows:

```{r eval=FALSE}
options(patRoon.MP.method = "future")

# set a future plan

# example 1: start a local cluster with four nodes
future::plan("cluster", workers = 4)

# example 2: start a networked cluster with four nodes on PC with hostname "otherpc"
future::plan("cluster", workers = rep("otherpc", 4)) 
```

Please see the documentation of the respective packages (_e.g._ [future] and [future.batchtools]) for more details on how to configure the workers.

The `withOpt()` function introduced in the previous subsection can also be used to temporarily switch between parallelization approaches, for instance:

```{r eval=FALSE}
# default to future parallelization
options(patRoon.MP.method = "future")
future::plan("cluster", workers = 4)

# ... do workflow

# do classic parallelization for GenForm
withOpt(MP.method = "classic", {
    formulas <- generateFormulas(fGroups, "genform", ...)
})

# .. do more workflow
```

#### Logging

Most tools that are executed in parallel will log their output to text files. These files may contain valuable information, for instance, when an error occurred. By default, the logfiles are stored in the `log` directory placed in the current working directory. However, you can change this location by setting the `patRoon.MP.logPath` option. If you set this option to `FALSE` then no logging occurs.

### Notes when using parallelization with futures

Some important notes when using the `future` parallelization method:

* `GenForm` currently performs less optimal with future multiprocessing to the `classic` approach. Nevertheless, it may still be interesting to use the `future` method to move the computations to another system to free up resources on your local system.
* Behind the scenes the [future.apply] package is used to schedule the tools to be executed. The `patRoon.MP.futureSched` option sets the value for the `future.scheduling` argument to the `future_lapply()` function, and therefore allows you to tweak the scheduling.
* Make sure that `patRoon` and in case of multiprocessing that the tool to be executed (`MetFrag`, `SIRIUS` etc.) are exactly the _same_ version on all computing hosts.
* Make sure that `patRoon` is properly configured on all hosts, _e.g._ set the `patRoon.path.XXX` options to ensure all tools can be found.
* For `MetFrag` annotation: if a local database such as `PubChemLite` is used, it must be present on each computing node as well. Furthermore, the local computer (even if not used for the computations) _also_ must have this file present. Like the previous point, make sure that the `patRoon.path.XXX` options are set properly.
* If you encounter errors then it may be handy to switch to `future::plan("sequential")` and see if it works or you get more descriptive error messages.
* In order to restart the nodes, for instance after re-configuring `patRoon`, updating `R` packages etc, simply re-execute `future::plan(...)`.
* Setting the `future.debug` package option to `TRUE` may give you more insight what is happening to find problems.


```{r child=file.path(vignDir, "shared", "_refs.Rmd")}
```
