# Ion mobility spectrometry (IMS-HRMS) workflows

## Introduction {#IMSWorkflow}

This chapter describes workflows to process data from Ion Mobility Spectrometry coupled to high resolution mass spectrometry (IMS-HRMS) instruments in `patRoon`. These are simply referred to as _IMS workflows_.

IMS is increasingly used to improve non-target analysis, and has the potential to improve separation of isomeric and isobaric compounds, clean up HRMS data and use the collision cross section (CCS) to improve identification of compounds. `patRoon 3.0` adds IMS support throughout the complete workflow and data processing functionality to take advantage of the additional information provided by IMS.

In `patRoon` three types of IMS workflows can be distinguished:

1. **LC-MS workflows with IMS data**: These workflows are like regular (non-IMS) workflows, but work with IMS data.
2. **Direct mobility assignment**: Feature mobilities are assigned directly during feature detection.
3. **Post mobility assignment**: Feature mobilities are assigned after finding and grouping features.

The following sections further detail each of these workflows, including the required data and the steps involved.

> **_NOTE_** IMS workflows are considerably more computationally demanding, as the raw IMS data typically contain several orders of magnitude more mass spectra. This is especially apparent during feature detection, but also subsequent steps such as creation of chromatograms and spectra will take up more time and RAM. The direct mobility assignment workflows are generally most demanding.

> **_NOTE_** While IMS workflows are largely the same as non-IMS workflows, there are some important considerations (outlined in this chapter). It is highly recommended to experiment with the [newProject tool](#newProject) and study the [example workflow](#exIMSWorkflow) to get a feeling of how IMS workflows work.

> **_NOTE_** IMS workflows have been primarily developed with Bruker TIMS and Agilent IMS data. Most of the functionality was designed to be generic and applicable to other IMS data as well, but data from other instruments likely need to be optimized. Furthermore, mobility and CCS conversion is not yet supported for other instruments. Contributions to test and extend support for other instruments are _very welcome!_

### LC-MS workflows with IMS data

In this workflow the raw IMS data is 'collapsed' and is converted in such a way it looks like 'regular' LC-MS data. This is achieved by summing up the mass spectra in each IMS frame and subsequently replacing the IMS frames by the combined spectra. The workflow then proceeds as a regular LC-MS workflow. Hence, this type of workflow does not bring any of the benefits of IMS data, but allows full compatibility with software tools that do not support IMS data. Furthermore, this type of workflow is much less computationally intensive, which makes it suitable for e.g. initial exploration of data.

### Direct mobility assignment {#directIMS}

This workflow relies on a feature detection algorithm that separates features by ion mobility and assigns the feature mobilities directly during feature detection. In `patRoon` this is currently only supported by the `piek` and `greedy` algorithms or by [importing feature data from other algorithms](#importFeat). Most of the remaining workflow is the same as non-IMS workflows. However, it typically adds steps to calculate CCS values for features and matches these with (predicted) CCS values of suspects and/or compound annotation candidates. Furthermore, the IMS data is internally used to cleanup extracted ion chromatograms and mass spectra, which can improve their visualization and quality of feature annotation.

### Post mobility assignment {#postIMS}

Post mobility assignment workflows can be considered as a hybrid approach to the two previous workflow types: the workflow starts with feature detection and grouping as in a regular LC-MS workflow, followed by assigning IMS data to features and then proceeding much like the direct mobility assignment workflow.

The mobility assignment consists of the following steps:

1. Feature detection and grouping is performed with classical algorithms like regular LC-MS workflows.
2. Extracted ion mobilograms are generated for each feature and automatic peak detection is used to detect mobilities.
3. Each of the detected mobilities for a feature are used to generate a new set of _IMS features_. These features inherit their properties from the original features, which are referred to as _IMS precursors_. A link is formed between the IMS features and the IMS precursor.
4. The IMS feature data such as retention time and intensities is updated from mobility filtered extracted ion chromatograms. Any features that could not be detected from the filtered data are removed.
5. The feature groups are updated with IMS data and similar links between _IMS feature groups_ and _IMS precursors_ are formed.

All these steps are automatically performed by the `assignMobilities()` method function that will be discussed later.

An advantage of post mobility workflows is the compatibility with any of the feature detection and grouping algorithms supported in `patRoon`. In addition, post assignment is often less computationally intensive than direct mobility assignment workflows. Furthermore, the links between IMS features and IMS precursors facilitate the recognition of possible protomers (i.e. redundant features of a same compound with the same retention time and _m/z_ but different mobility) and assists in recognizing IMS features across MS polarities in [sets workflows](#setsWorkflow). Finally, post mobility assignment workflows can selectively fallback to 'regular' LC-MS feature data from the IMS precursors in steps where this makes more sense, this will be discussed later. Potential disadvantages of post mobility assignment workflows are the reliance on raw data that is with and without IMS dimension, and the mobility assignment typically relies on two steps of peak detection (steps 2 and 4) which may fail with e.g. low intensity data. Nevertheless, this type of workflow is usually recommended for most IMS data processing projects.

### Summary

The following table summarizes the differences between the three IMS workflows (some of these are introduced later in this chapter):

|                               | LC-MS with IMS data | Direct mobility assignment | Post mobility assignment
------------------------------- | ------------------- | -------------------------- | --------------------------
Raw data requirements^1^        | IMS collapsed       | Raw IMS data               | IMS and IMS collapsed raw data
Feature detection and grouping  | Regular non-IMS (all algorithms) | Utilizes IMS separation (few algorithms) | Regular non-IMS (all algorithms)
Feature mobility assignment     | No                  | During feature detection   | After feature grouping
Feature CCS assignment          | No                  | Yes                        | Yes
IMS precursor links             | No                  | No                         | Yes
Suspect & compound CCS matching | No                  | Yes                        | Yes
HRMS cleanup                    | No                  | Yes                        | Yes
non-IMS feature fallback        | -                   | No                         | Yes
Supports [sets workflows](#setsWorkflow)^2^ | Yes     | Limited                    | Moderate

Notes:

1. If the `piek` feature detection algorithm is used then (only) IMS raw data may suffice, see [more info here](#fileTypes).
2. Discussed further [here](#IMSSetsWorkflow) 

## Performing IMS workflows

### Parameter defaults {#limitsIMS}

To effectively process IMS-HRMS data, it may be required to adjust the default numerical limits (e.g. mobility tolerances) and other settings. This is especially important since (1) different IMS instruments typically use different mobility units and (2) the data is often more 'raw', i.e. more noisy and without centroiding, which therefore require more instrument sensitive optimization. To simplify this, `patRoon` brings defaults for Bruker timsTOF and Agilent IMS instruments. The default instrument type and general numeric limits, are specified in the [limits](#limits) file. The instrument type that is defined there, also affects the default settings used for `piek`, mobilogram extraction and peak list generation (discussed later). To generate a limits file with the default settings for Agilent IMS data, the following code can be used:

```{r exLimitsIMS, eval=FALSE}
genLimitsFile(IMS = "agilent")
```

A limits file is normally generated when using the `newProject()` function to create a new project. If no limits file is present, the defaults are used from Bruker instruments. See the [limits](#limits) advanced section for more details on the limits file and how to adjust it.

### Raw data

The [data conversion and pre-treatment](#pre-treat) section already discussed how raw data can be converted to make it suitable for processing. For the workflows that require raw data that has IMS data (see the Table of the previous section) there are two options:

1. Use the raw instrument directly. This is currently only supported for Bruker TIMS data, see [details here](#fileTypes).
2. Convert the data to the `ims` type in the `mzML` file format, e.g. by using `convertMSFiles()` ([explained here](#pre-treat)).

If IMS collapsed data is needed (see the Table of the previous section) then raw data of the type `raw` or `ims` should be converted to the type `centroid`.

The [data conversion and pre-treatment](#pre-treat) section includes some examples on how to do these conversion steps.

### Direct mobility assignment (DMA) workflows

#### Feature detection 

In direct mobility assignment (DMA) workflows the feature detection needs to be aware of the IMS dimension. In `patRoon` this is currently only supported by the `piek` algorithm. Alternatively, feature data from another IMS aware feature detection [can also be imported](#importFeat).

The `piek` algorithm was already [discussed before](#piek). By default, `piek` does not use IMS data for feature
detection, even if processing IMS-HRMS data. To enable the use of IMS data, the `IMS` function argument be set to
`TRUE`. For instance:

```{r exDMA, eval=FALSE}
# comprehensive feature detection from 2 dimensional EIC bins
genEICParams <- getPiekEICParams(mzRange = c(100, 800), mobRange = c(0.5, 1.5), mzStep = 0.02, mobStep = 0.2)
peakParams <- getDefPeakParams("chrom", "piek")
fList <- findFeatures(anaInfo, "piek", IMS = TRUE, genEICParams = genEICParams, peakParams = peakParams)
```

##### Optimizing computational demands

When `piek` is used to detect IMS resolved features, it will generate features from EIC bins that are created from the _m/z_ **and** mobility dimension. This can result in a very large number of EIC bins, which can make feature detection computationally intensive. For this reason, it is recommended to have sufficient amount of RAM (e.g. >= 32GB).

In addition, there are several advanced EIC parameters that can be tweaked to reduce computational demands (see `?getPiekEICParams` for a complete overview). For instance:

EIC parameter     | Description
----------------- | ------------------------------------------------
`mzRange`, `mobRange` | Limits the dimensions of the EIC bins. Reducing this can significantly speed up feature detection.
`retRange`        | Limits the retention time range of the EIC bins. Useful to e.g. discard features from the dead volume.
`filter`, `filterIMS` | Applies filters which can considerably reduce the number of EIC bins. The `filter` parameter [was discussed before](#piek). The `filterIMS` parameter is analogous to `filter`, but applies the given filter in the mobility dimension. `filterIMS` should be either `"none"` for no mobility filtering or equalling `"filter"` (i.e. `"suspects"` or `"ms2"`).
`minEICIntensity` | Discard EICs with a maximum intensity below the given intensity threshold.
`minEICAdjTime`, `minEICAdjPoints`, `minEICAdjIntensity` | Discard EICs without a clear continious signal.
`topMostEICMob`   | Only keep these top most intense EICs (based on maximum signal intensity).

Furthermore, the `EICBatchSize` function argument can be set to split the work in multiple batches, which is especially useful to reduce RAM usage. The following examples show how these parameters can be used to reduce computational needs:

```{r exDMAOpt, eval=FALSE}
# filter EIC bins by m/z and mobility data from detected PASEF MS2 precursors
genEICParams <- getPiekEICParams(filter = "ms2", filterIMS = "ms2", minTIC = 1000)
fList <- findFeatures(anaInfo, "piek", IMS = TRUE, genEICParams = genEICParams)

# filter by suspect m/z data, but don't filter mobility bins (e.g. if suspect list doesn't contain IMS data)
genEICParams <- getPiekEICParams(filter = "suspects", filterIMS = "none", mobRange = c(0.5, 1.5))
fList <- findFeatures(anaInfo, "piek", IMS = TRUE, genEICParams = genEICParams,
                      suspects = suspList, adduct = "[M+H]+")

# focus on low m/z and mobility features with high intensity
# don't load and process more than 10,000 EICs at a time
genEICParams <- getPiekEICParams(mzRange = c(100, 300), mobRange = c(0.5, 0.8), minEICIntensity = 1E5)
fList <- findFeatures(anaInfo, "piek", IMS = TRUE, genEICParams = genEICParams, EICBatchSize = 1E4)
```

##### m/z and mobility determination

The raw _m/z_ and mobility data is typically not or partially centroided. For this reason, `piek` automatically generates _m/z_ and mobility _versus_ intensity peak profiles for each feature, and determines the _m/z_ and mobility data from these profiles. This process is largely similar to centroiding in HRMS workflows. For accurate _m/z_ and mobility determination, it may be needed to tweak the default parameters that are used for profile generation:

EIC profile parameter | Description
--------------------- | ------------------------------------------------
`sumWindowMZ`, `sumWindowMob` | The retention window (+/- seconds) for summing up raw data to generate the profiles. Summing up multiple data points is especially needed for noisy data, e.g. generated with fast timsTOF methods. It's best to keep this smaller than the expected minimum LC peak width to avoid summing up data from different chromatographic peaks. Setting this to zero will dsiable summing.
`smoothWindowMZ`, `smoothWindowMob`, `smoothExtMZ`, `smoothExtMob` | Smoothing parameters. This is again primarily useful for noisy data. Setting these to zero will disable smoothing.

##### Instrument defaults

To simplify the configuration of the previously discussed parameters, `piek` brings defaults for two different instrument types. The instrument is set by the `IMS` argument for `getPiekEICParams()` and should be set to `"bruker"` or `"agilent"`, for Bruker timsTOF and Agilent IMS data, respectively. The default value for `IMS` is determined from the value determined from the previously discussed [limits file](#limitsIMS).


#### Feature grouping

The grouping of features in DMA workflows is currently only possible with the `greedy` algorithm or by [importing data](#importFeat). The `greedy` algorithm was already [briefly introduced before](#groupFeatures). This algorithm was introduced in `patRoon 3.0` and is a simple and fast grouping algorithm, which also supports IMS data. Its `mobWindow` argument can be set to configure the mobility grouping tolerance. Some examples:

```{r exGroupGreedyIMS, eval=FALSE}
fGroups <- groupFeatures(fList, "greedy", mobWindow = 1)

# customize all tolerances and weights
fGroups <- groupFeatures(fList, "greedy",
                         rtWindow = 5,
                         mzWindow = 0.003,
                         mobWindow = 0.5,
                         scoreWeights = c(retention = 1, mz = 3, mobility = 3, intensity = 1))
```

### Post mobility assignment (PMA) workflows

#### Feature detection and sample grouping

In post mobility assignment (PMA) workflows, the feature detection and grouping is performed as in regular LC-MS workflows, i.e. without using the IMS data. Hence, the `findFeatures()` and `groupFeatures()` functions are used exactly like what was discussed before [for HRMS workflows](#features).

#### Assigning mobility values to features {#assignMobFeat}

The `assignMobilities()` method function is used to assign mobilities to features. The most important function arguments are:

Argument                 | Description
------------------------ | ------------------------------------------------
`mobPeaksParams`, `chromPeaksParams` | Parameters for the peak detection in mobilograms and chromatograms, respectively. See `?getDefPeakParams()` for more details.
`EIMParams`, `EICParams` | Parameters for the extraction of ion mobilograms (EIMs) and chromatograms (EICs), respectively. See `?getDefEIMParams()` and `?getDefEICParams()` for more details.
`fallbackEIC`            | Set to `TRUE` to use the raw EIC data to update properties of IMS features, in case peak detection fails in mobility filtered EICs ([steps 3-4](#postIMS)).
`fromSuspects`           | _Suspect screening workflows_: Use suspect data for mobility assignment, this is [discussed later](#assignMobFromSusps).
`IMSMatchParams`         | _Suspect screening workflows_: parameters to match IMS data between features and suspects, this is [discussed later](#screenIMS).


Note that mobility assignment is only performed if either `mobPeaksParams` or `fromSuspects` is set. The latter is [discussed later](#assignMobFromSusps). The `assignMobilities()` method function is executed after executing `groupFeatures()`, with often some intermediate steps in-between (e.g. filtering, componentization, discussed later in this chapter). A single call to `assignMobilities()` will perform all [the mobility assignment steps that were introduced before](#postIMS). 

The `EIMParams` parameter influences the construction of extracted ion mobilograms (EIMs), which are used to determine the feature mobilities. Importantly, for noisy data (e.g. obtained with fast timsTOF methods), the enabling if smoothing may be needed. The `IMS` argument to `getDefEIMParams()` can be set to `"bruker"` or `"agilent"` to use defaults for Bruker timsTOF and Agilent IMS data, respectively. The default value for `IMS` is determined from the value determined from the previously discussed [limits file](#limitsIMS).

Some examples of how to use `assignMobilities()` for mobility assignment are given below:

```{r exPostMob, eval=FALSE}
fGroups <- assignMobilities(fGroups, mobPeaksParams = getDefPeakParams("bruker_ims", "piek"),
                            chromPeaksParams = getDefPeakParams("chrom", "piek"))

# customized peak detection parameters
fGroups <- assignMobilities(fGroups, mobPeaksParams = getDefPeakParams("agilent_ims", "xcms3", peakwidth = c(0.2, 3)),
                            chromPeaksParams = getDefPeakParams("chrom", "openms", gaussWidth = 20))

# use (aggressive) smoothing for noisy EIM data to improve mobility determination
fGroups <- assignMobilities(fGroups, mobPeaksParams = getDefPeakParams("bruker_ims", "piek"),
                            chromPeaksParams = getDefPeakParams("chrom", "piek"),
                            EIMParams = getDefEIMParams(smooth = "sg", smLength = 25))

# like above, setting changing instrument defaults for Agilent data
fGroups <- assignMobilities(fGroups, ..., EIMParams = getDefEIMParams(IMS = "agilent"))
```

Further details are found in the reference manual (`?assignMobilities_feat`).

### Mobility and CCS conversion {#mobCCSConv}

Conversion of mobility and CCS values in DMA and PMA workflows is performed with the `assignMobilities()` function introduced in the previous section, and is activated if its `CCSParams` function argument is set. This argument should be assigned with a `list` that contains the parameters to configure the conversions process. The parameters are instrument specific and the parameter list can be obtained with the `getCCSParams()` function. The most important arguments for `getCCSParams()` are:

Function argument | Description
----------------- | ---------------------------------------------------------------
`method`          | Specifies the conversion method. Valid values are: `"bruker"`, `"mason-schamp_k"` `"mason-schamp_1/k"` or `"agilent"`.
`calibrant`       | Specifies the calibrant file or parameters if `method="agilent"`.

The function currently only supports Bruker TIMS and Agilent IMS data (contributions to Waters data and information are _very welcome!_). For Bruker TIMS instruments, the `method` should be set to `"bruker"` or `"mason-schamp_1/k"`. The former relies on the Bruker TDF-SDK (see the [Installation](#Installation) chapter to install it). The latter doesn't, but may give very minor differences in the results compared to how Bruker software performs its calculations. For Agilent IMS data, the `method` should be set to `"agilent"` and the `calibrant` argument should be set to configure the calibration settings. The value for `calibrant` is either a file path to the `.d` or `OverrideImsCal.xml` file that should be used for re-calibration, or a `list` with the elements `massGas`, `TFix` and `beta`. It is recommended to read the reference manual for the `getCCSParams()` function for more details (`?getCCSParams`).

Some examples of how to use `assignMobilities()` for mobility and CCS conversion are given below:

```{r exMobCCS, eval=FALSE}
# conversion of Bruker TIMS data using the TDF-SDK
fGroups <- assignMobilities(fGroups, CCSParams = getCCSParams("bruker"))

# conversion of Agilent IMS data
fGroups <- assignMobilities(fGroups, CCSParams = getCCSParams("agilent", calibrant = "path/to/Calibrant.d"))

# combine mobility assignment and CCS conversion into one step
fGroups <- assignMobilities(fGroups, mobPeaksParams = getDefPeakParams("bruker_ims", "piek"),
                            chromPeaksParams = getDefPeakParams("chrom", "piek"),
                            CCSParams = getCCSParams("bruker"))
```

### Suspect screening {#IMSSuspScr}

[Suspect screening](#suspscr) in IMS workflows is mostly the same as in non-IMS workflows. However, the suspect list can be extended with IMS properties to improve suspect matching, and performing the screening step may be slightly different for PMA workflows.

#### Suspect list

The suspect list can contain the following additional columns (all optional):

* `mobility`: The mobility of the suspect.
* `CCS`: The collision cross section (CCS) of the suspect.
* `mobility_<adduct>`, `CCS_<adduct>`: The mobility and CCS of the suspect for a specific adduct. The `<adduct>` should be replaced with the adduct name, e.g. `mobility_[H+Na]+` or `CCS_[M-H]-`.

Multiple suspect reference values can specified in each of the columns by separating them with a semicolon (`;`). If data for both the adduct and non-adduct specific columns is available (and not `NA`), then the latter gets precedence. It is recommended to include the adduct specific data whenever possible. During the suspect screening step to correct data is automatically chosen based on the adduct assigned to the feature (or assigned to the `adduct` function argument).

#### Prediction of mobility and CCS data {#assignMobSusps}

The `assignMobilities()` function that was discussed before also has a method specifically for suspect lists. It can be used for the prediction (or library matching) of IMS data for suspects and convert CCS values to mobilities (or vice versa). The common function arguments include:

Function argument | Description
----------------- | ---------------------------------------------------------------
`from`            | Specifies from where IMS data is added. See below for more details.
`adducts`         | Specifies for which adducts IMS data is added or converted. A `character` vector with multiple values is allowed. Include `NA` to also consider the usage of non-adduct specific data. Any adducts specified in the suspect list (`adduct` column) are always considered.
`CCSParams`       | Parameters for the mobility and CCS conversions. See [its previous introduction](#mobCCSConv).
`overwrite`       | Set to `TRUE` to overwrite any existing IMS data in the suspect list.

The following options exist for the `from` function argument:

* `from=NULL` (default): No IMS data is added (but mobility and CCS conversion may still occur).
* `from="pubchemlite"`: IMS data is matched from the [PubChemLite database with CCS][PCLiteCCS-dl] data. See the [Installation chapter](#Installation) for details on how to install the database.
* `from="c3sdb"`: Uses [C3SDB] to predict CCS values for suspects. See the [Installation chapter](#Installation) for installation details.
* a `data.table` or `data.frame`: A custom library with IMS data.

Some examples:

```{r assignMobSusps, eval=runData}
suspList <- patRoonData::suspectsPos[1:4, ]

# adds CCS values from PubChemLite and converts them to mobilities
suspList <- assignMobilities(suspList, from = "pubchemlite", adducts = c("[M+H]+", "[M+Na]+", "[M+H]-"),
                             CCSParams = getCCSParams("mason-schamp_1/k"))
suspList[, grepl("name|InChIKey|mobility|CCS", colnames(suspList))]

# overwrites data from custom library
IMSLib <- data.frame(
    name = c("DEET", "Diglyme"), # (name column is not mandatory)
    InChIKey1 = c("MMOXZBCLCQITDF", "SBZXBUIDTXKZTM"), # NOTE: matching happens by first-block IK by default
    "CCS_[M+H]+" = c(150, 160),
    check.names = FALSE # NOTE: need this to allow special characters in column names (ie in CCS column)
)
suspList <- assignMobilities(suspList, from = IMSLib, overwrite = TRUE,
                             adducts = "[M+H]+", CCSParams = getCCSParams("mason-schamp_1/k"))
suspList[, grepl("name|InChIKey|mobility|CCS", colnames(suspList))]
```

#### Performing suspect screening {#screenIMS}

The `screenSuspects()` method function [introduced before](#suspscr) is also used to perform the suspect screening in IMS workflows. The `IMSMatchParams` argument configures how IMS data is used to match suspects and features. See `?getIMSMatchParams` and the examples below for more details.

In post mobility assignment workflows, it is possible to perform the suspect screening _before_ `assignMobilities()` is executed to assign the feature mobilities. In this case the `IMSMatchParams` argument can also be passed to `assignMobilities()` to filter out any suspect hits with deviating IMS data after feature mobilities have been assigned. This may be useful to e.g. first prioritize features with suspect screening and other steps before assigning mobilities to the remaining features.

Some examples are shown below:

```{r screenIMS,eval=FALSE}
# suspList is a suspect list with CCS and mobility data (see previous subsections)

# screen with default CCS matching parameters
fGroupsScr <- screenSuspects(fGroups, suspects = suspList, adduct = "[M+H]+",
                             IMSMatchParams = getIMSMatchParams("CCS"))

# match suspects on mobility with a +/- 0.1 tolerance
fGroupsScr <- screenSuspects(fGroups, suspects = suspList, adduct = "[M+H]+",
                             IMSMatchParams = getIMSMatchParams("mobility", window = 0.1, relative = FALSE))

# match suspects on CCS with a +/- 6% tolerance
fGroupsScr <- screenSuspects(fGroups, suspects = suspList, adduct = "[M+H]+",
                             IMSMatchParams = getIMSMatchParams("CCS", window = 0.06, relative = TRUE))

# post mobility assignment workflow with prior suspect screening
# regular non-IMS suspect screening
fGroupsScr <- screenSuspects(fGroups, suspects = suspList, adduct = "[M+H]+", onlyHits = TRUE)
# ... do other things such as more feature prioritization
# ... and assign mobilities to the remaining features, match already screened suspects by CCS
fGroupsScr <- assignMobilities(fGroupsScr, mobPeaksParams = getDefPeakParams("bruker_ims", "piek"),
                               chromPeaksParams = getDefPeakParams("chrom", "piek"),
                               IMSMatchParams = getIMSMatchParams("CCS"))
```

#### Assigning feature mobilities from suspect data {#assignMobFromSusps}

In post mobility assignment workflows, the `assignMobilities()` method function can also be used to assign feature mobilities from suspect data by setting the `fromSuspects` argument to `TRUE`. This replaces the mobility assignment step from mobilograms ([step 2](#postIMS)), and copies the assigned reference value for a suspect directly to feature data. If both `fromSuspects` and `mobPeaksParams` are set, then the mobility detection from mobilograms is performed for features without a suspect hit. If a feature has multiple hits, it is unclear which suspect value should be used, and therefore _no_ suspect data will be used for mobility assignment in this case. The `IMSRange` function argument is used to derive the mobility range for the feature. Assigning mobilities from suspect data is only supported if the suspect screening was already performed prior to calling `assignMobilities()`.

Bypassing the need for peak detection for mobility assignment may be advantageous for low intensity features. However, the mobility values assigned to features are not verified against the raw data. The feature mobility boundaries are also not derived from experimental data, hence, the assigned mobility range to the feature may not cover the actual experimental range (i.e. the peak width in a mobilogram). Thus, when intensities are updated for IMS features ([step 4](#postIMS)), these can be superficially low (i.e. partial removal of signal) or high (i.e. inclusion of data of neighboring mobility peaks). With these limitations in mind, this approach is primarily intended for the following scenarios:

1. The mobility for a suspect is accurately known in advance, and no additional peak detection is needed. In this case the `mobWindow` argument is typically set to a narrow tolerance window.
2. There is no interest for accurate mobility (and CCS) assignment, and IMS data should only be used as a rough filtering step. In this scenario the `mobWindow` argument is typically set to a relatively large tolerance window.

Below are some example of how to use `assignMobilities()` to assign feature mobilities from suspect data:

```{r assignMobFromSusps, eval=FALSE}
# suspect screening must be performed in advance
# suspList is a suspect list with IMS data (see previous subsections)
fGroups <- screenSuspects(fGroups, suspects = suspList, adduct = "[M+H]+", onlyHits = TRUE)

# scenario #1: assume suspect data is accurate
fGroups <- assignMobilities(fGroups, fromSuspects = TRUE, mobWindow = 0.01,
                            chromPeakParams = getDefPeakParams("chrom", "piek"))

# scenario #2: wide tolerance window, no interest in accurate mobility assignment
# add fallback to mobility detection for features with no (or >1) suspect hit or if no suspect mobility data is available
# set IMSMatchParams to also filter out suspect hits with deviating mobilities
fGroups <- assignMobilities(fGroups, fromSuspects = TRUE, mobWindow = 0.1,
                            mobPeaksParams = getDefPeakParams("bruker_ims", "piek"),
                            chromPeakParams = getDefPeakParams("chrom", "piek"),
                            IMSMatchParams = getIMSMatchParams("CCS"))
```

For more details see the reference manual (`?assignMobilities_feat`).

### Componentization

The algorithms that use componentization to detect adducts, isotopes, homologous series etc. (e.g. [CAMERA], [cliqueMS] and [nontarget]) currently do not support IMS data. They are not supported in direct mobility assignment workflows. However, in post mobility assignment workflows they can operate on the IMS precursors (either before or after calling `assignMobilities()`).

The other componentization algorithms (e.g. intensity clustering and [transformation products](#TPs)) optionally include IMS features and IMS precursors during the componentization. This is controlled by the `IMS` function argument to the `generateComponents()` function, and can be set as following:

1. `IMS=FALSE`: Do not consider any IMS features (only supported in post mobility assignment workflows).
2. `IMS=TRUE`: Only consider IMS features and ignore any IMS precursors.
3. `IMS="both"`: Consider both IMS features and IMS precursors.
4. `IMS="maybe"`: Consider IMS precursors if present, and IMS features otherwise. This is the default value.

In the case that IMS features are not considered (i.e. `IMS=FALSE` or `IMS="maybe"`), then the componentization data from the IMS precursors is directly copied to the IMS features. This is useful when the componentization results for IMS precursors and IMS features are expected to be similar, which is often the case for the algorithms that support the `IMS` argument.

In post mobility assignment workflows, it is possible to perform the componentization prior to calling `assignMobilities()`. In this case the componentization results can be 'expanded' by copying the IMS precursor results to the IMS features afterwards. This is performed by the `expandForIMS()` method function:

```{r expandComponIMS, eval=FALSE}
# perform componentization
components <- generateComponents(fGroups, "intclust")
# assign mobilities
fGroups <- assignMobilities(fGroups, mobPeaksParams = getDefPeakParams("bruker_ims", "piek"),
                            chromPeaksParams = getDefPeakParams("chrom", "piek"))
# expand componentization results for IMS features
components <- expandForIMS(components, fGroups)
```

### Annotation

IMS workflows bring several advantages to feature annotation. The `generateMSPeakLists()` function automatically filters MS and MS/MS spectra by the mobility range assigned to the feature, which can considerably improve the quality of peak lists. Furthermore, CCS data can be assigned to compound annotation candidates and used to eliminate unlikely candidates.

> **_NOTE_** In post mobility assignment workflows it is important that the `assignMobilities()` method function is called _prior_ to any feature annotation steps in order to take advantage IMS data during feature annotation.

#### MS peak lists

The `generateMSPeakLists()` function automatically filters MS and MS/MS spectra by mobility data assigned to features, and is therefore used similarly like non-IMS workflows. The function supports Bruker PASEF experiments, and can therefore assign _m/z_ and mobility filtered MS/MS data to features. In post mobility assignment workflows, the MS and MS/MS data that will be assigned to IMS precursors are based on combined spectra for each IMS frame.

The _m/z_ values in IMS-HRMS data are typically not (e.g. Agilent) or partially (e.g. Bruker) centroided. Since feature annotation algorithms are usually limited to centroided data, `generateMSPeakLists()` performs an additional centroiding step for IMS-HRMS data. The centroiding parameters, and some other advanced IMS specific parameters, are configured through the averaging parameters given to `generateMSPeakLists()` (`avgFeatParams` arguments). The default parameters can be changed with the `IMS` argument to `getDefAvgPListParams()`, which can be set to `"bruker"` or `"agilent"` for Bruker and Agilent IMS data, respectively. The default value for `IMS` is determined from the value determined from the previously discussed [limits file](#limitsIMS). For example:

```{r exAvgPListIMS, eval=FALSE}
# generate peak lists with default parameters for Agilent data
mslists <- generateMSPeakLists(fGroups, avgFeatParams = getDefAvgPListParams(IMS = "agilent"))
```

see the reference manual for more details (`?getDefAvgPListParams`).

#### Compound annotation

Candidates from compound annotations can be matched with IMS data for features, similarly as was discussed for [candidates from suspect screening](#IMSSuspScr). The `assignMobilities()` method function works similarly to assign IMS data to candidates and perform mobility and CCS conversions. The `filter()` method function can subsequently be used to eliminate candidates that match poorly with features ([discussed later](#IMSFiltering)).

The `assignMobilities()` function has similar arguments as the method for suspect screening:

Function argument | Description
----------------- | ---------------------------------------------------------------
`fGroups`         | The feature groups for which the `compounds` were generated.
`from`            | Specifies from where IMS data is added. See [the suspects section](#assignMobSusps) for more details.
`adduct`          | Should match the value of the `adduct` argument that was passed to `generateCompounds()`. Set to `NULL` if [adduct annotations are available for features](#incorpAdductIso) or in [sets workflows](#setsWorkflow).
`CCSParams`       | Parameters for the mobility and CCS conversions. See [the previous discussion](#mobCCSConv).
`overwrite`       | Set to `TRUE` to overwrite any existing IMS data in the candidate list.
`IMS`             | For which type of features IMS data should be added to the candidates: `FALSE` (only IMS precursors), `TRUE` (only IMS features, default), `"both"` (both IMS precursors and IMS features) or `"maybe"` (IMS precursors if available, otherwise IMS features).

> **_NOTE_** If [MetFrag] with the [PubChemLite database with CCS][PCLiteCCS-dl] is used for compound annotation, then the CCS data is copied from the database and there is no need to set the `from` argument.

Some examples:

```{r exCompIMS, eval=FALSE}
# generate compound candidates
compounds <- generateCompounds(fGroups, mslists, "metfrag", adduct = "[M+H]+", database = "comptox")
# perform CCS prediction with C3SDB and convert the values to mobilities
compounds <- assignMobilities(compounds, fGroups, from = "c3sdb", adduct = "[M+H]+",
                              CCSParams = getCCSParams("bruker"))

# generate compound candidates
# NOTE: if PubChemLite with CCS data is installed, then CCS data is automatically added
compounds <- generateCompounds(fGroups, mslists, "metfrag", adduct = "[M+H]+", database = "pubchemlite")
# overwrite CCS data from a custom library (candidates not in the library will be untouched)
IMSLib <- data.frame(
    name = c("Caffeine", "Acetaminophen"), # name column is not mandatory
    InChIKey = c("RYYVLZVUVIJVGH-UHFFFAOYSA-N", "RZVAJINKPMORJF-UHFFFAOYSA-N"),
    CCS = c(150, 160)
))
compounds <- assignMobilities(compounds, fGroups, from = IMSLib, adduct = "[M+H]+",
                              overwrite = TRUE)

```

For more details see the reference manual (`?assignMobilities_comp`).

#### Copying annotation data to IMS features with similar MS/MS data

The number of features in post mobility assignment workflows can be very large, considering that both IMS features and their IMS precursors are present. Since formula and compound annotation is primarily influenced by MS/MS data, it is possible to skip annotation for IMS features with highly similar MS/MS data as their IMS precursor, and copy the annotations from the IMS precursor result instead. The `minIMSSpecSim` argument to `generateFormulas()` and `generateCompounds()` sets the minimum spectral similarity threshold for when MS/MS data is considered similar.

```{r minIMSSpecSim, eval=FALSE}
# generate formulas: copy IMS precursor results when the spectral similarity is at least 0.9
formulas <- generateFormulas(fGroups, mslists, "genform", adduct = "[M+H]+", minIMSSpecSim = 0.9)
```

### Sets workflows {#IMSSetsWorkflow}

There is basic functionality to combine IMS workflows and [sets workflows](#setsWorfklow). However, an important limitation currently is that there is no way to group IMS features across the different sets. This is because mobility values among polarities (e.g. `[M+H]+` and `[M-H]-`) typically differ slightly and there is not a straightforward way to normalize for this. However, in post mobility assignment workflows grouping of IMS precursors across sets still occurs (similar to non-IMS workflows). Hence, the links between IMS features and IMS precursors can still give some indication which IMS features across sets are related. Thus, combined IMS and sets workflows are therefore mainly useful to e.g. process positive and negative sample analyses in one workflow, but lack other advantages of non-IMS sets workflows, such as easily relating features among sets and improving feature annotation with combined positive/negative MS/MS data.

Like non-IMS workflows, the `makeSet()` function is used to initialize the sets workflow.

## Processing data

```{r include=FALSE,eval=runData}
fGroupsO <- fGroups
fGroups <- fGroupsIMS
mslistsO <- mslists
mslists <- mslistsIMS
compoundsO <- compounds
compounds <- compoundsIMS
```

The data processing functionality of `patRoon` that was [discussed before](#processing) equally applies to IMS workflows. In addition, some extra functionality exists to inspect and filter the data. These are discussed in the next subsections.

### Updating feature group properties

The `updateGroups()` function to update feature group properties (retention time and _m/z_) was [briefly introduced before](#updFG). This function can also be used to update the assigned mobilities and CCS values to e.g. improve their accuracy after eliminating unwanted features:

```{r exUpdFGIMS, eval=FALSE}
fGroups <- filter(fGroups, ...) # perform filtering steps

fGroups <- updateGroups(fGroups) # update all by default: retention time, m/z, mobility and CCS (if present)
fGroups <- updateGroups(fGroups, what = c("mobility", "CCS")) # only update mobility and CCS
```

### Inspecting and plotting data

The `plotMobilograms()` method function is used to plot extracted ion mobilograms (EIMs), and works very similar as the `plotChroms()` function [introduced before](#vis_feat_ann). The function arguments are mostly equal, and the `EIMParams` argument is similarly used to configure advanced parameters (see `?getDefEIMParams` for details).

Most other plotting functions have an `IMS` argument that controls the inclusion of IMS features and IMS precursors in the plots. This argument is similar as discussed before for e.g. componentization and feature annotation and supports the following values:

- `IMS=FALSE`: Do not plot any IMS features (only supported in post mobility assignment workflows).
- `IMS=TRUE`: Only plot IMS features and ignore any IMS precursors.
- `IMS="both"`: Plot both IMS features and IMS precursors.
- `IMS="maybe"`: Plot IMS precursors if present, and IMS features otherwise. This is the default value.

The default is `"maybe"` since most plotting functions primarily use LC-MS data and therefore produce similar plots for IMS features and their IMS precursors.

The `spectrumSimilarityIMS()` method function calculates the spectrum similarity between MS (or MS/MS) data of a IMS feature and its IMS precursor. This function is internally used by feature annotation if the `minIMSSpecSim` argument is set, and uses the `spectrumSimilarity()` function [introduced earlier](#specSim).

The [reporting functionality](#report) of `patRoon` contains various functionality that facilitates processing of IMS data.

Some examples are shown below:

```{r exInspPlotIMS, eval=runData}
# plot mobilogram for first IMS precursor
plotMobilograms(fGroups[1, "M146_R301_22"])
# plot mobilogram for an IMS feature: this highlights the mobility (range) assigned to the feature
# NOTE: the mobility is appended to the feature group name
plotMobilograms(fGroups[1, "M146_R301_22_I0.57"], showPeakArea = TRUE, showFGroupRect = FALSE)

plot(fGroups, IMS = TRUE) # make a scatter plot with only IMS features

# returns a table with spectral similarities between IMS features and their IMS precursors
spectrumSimilarityIMS(mslists, fGroups)
```

### Subsetting and filtering data {#IMSFiltering}

The following filters are specifically for IMS workflows available:

Filter             | Classes         | Remarks
-------------------|-----------------|---------------------------------------------------------------------------
`IMS`              | `featureGroups` | Only keep IMS precursors (`IMS=FALSE`), IMS features (`IMS=TRUE`), both (`IMS="both"`) or IMS precursors if available and IMS features otherwise (`IMS="maybe"`). The `IMS` argument is also available for the `[` operator.
`withIMSPrecursor` | `featureGroups` | Only keep IMS features that have an IMS precursor.
`applyIMS`         | `featureGroups` | Only apply other filters to a subset of features. Should be set like the `IMS` argument.
`IMSRangeParams`   | `features`, `featureGroups`, `compounds` | Only keep features/annotations within the specified mobility range. The range is configured with the `getIMSRangeParams()` function.
`IMSMatchParams`   | `compounds`     | Only keep candidates that match reference IMS data. See [its description in suspect screening workflows](#screenIMS) for more details.

Some examples are shown below:

```{r exFiltIMS, eval=FALSE}
# keep only IMS features
# NOTE: this effectively results in data that is equal to a direct IMS workflow
fGroupsMob <- filter(fGroups, IMS = TRUE)
fGroupsMob <- fGroups[IMS = TRUE] # same as above

# only keep IMS precursors (post mobility assignment workflows only)
# NOTE: this effectively results in data that is equal to before `assignMobilities()` was called
fGroupsIMSPar <- filter(fGroups, IMS = FALSE)

# perform various filtering steps and ensure that only IMS features with an IMS precursor are kept
fGroupsF <- filter(fGroups, absMinIntensity = 1E5, relMinReplicateAbundance = 1,
                   withIMSPrecursor = TRUE)

# perform various filtering steps, but not on IMS features
fGroupsF <- filter(fGroups, absMinIntensity = 1E5, relMinReplicateAbundance = 1,
                   applyIMS = FALSE)

# only keep compound candidates within a defined CCS range
compoundsF <- filter(compounds, IMSRangeParams = getIMSRangeParams("CCS", 150, 200))

# only keep compound candidates with matching CCS (6% tolerance)
compoundsF <- filter(compounds, IMSMatchParams = getIMSMatchParams("CCS", window = 0.06, relative = TRUE))
```


```{r include=FALSE,eval=runData}
fGroups <- fGroupsO
mslists <- mslistsO
compounds <- compoundsO
```

## Example workflow {#exIMSWorkflow}

The example below shows a complete IMS workflow which uses post mobility assignment for features, and includes matching of suspect and compound annotation candidates by CCS.

> **_NOTE_** The [newProject tool](#newProject) tool supports the creation of IMS workflows, and makes it easy to to create and explore different types of IMS workflows.

```{r exIMSWorkflow, eval=FALSE}
# anaInfo should be a data.frame with the analysis information

# convert raw data to ims and to centroid data
convertMSFiles(anaInfo, typeFrom = "raw", formatFrom = "bruker_ims", typeTo = "ims", formatTo = "mzML",
               algorithm = "pwiz")
convertMSFiles(anaInfo, typeFrom = "ims", formatFrom = "mzML", typeTo = "centroid", formatTo = "mzML",
               algorithm = "imscollapse")

# perform feature detection and grouping
fList <- findFeatures(anaInfo, "openms")
fGroups <- groupFeatures(fList, "openms")

# perform basic filtering
fGroups <- filter(fGroups, absMinIntensity = 1E4, relMinReplicateAbundance = 1)

# assign mobilities and CCS values
CCSParams <- getCCSParams("mason-schamp_1/k")
fGroups <- assignMobilities(fGroups, mobPeaksParams = getDefPeakParams("bruker_ims", "piek"),
                            chromPeaksParams = getDefPeakParams("chrom", "piek"),
                            CCSParams = CCSParams)

# suspect screening
suspList <- read.csv("suspects.csv")
# add CCS data from PubChemLite and convert them to mobilities
suspList <- assignMobilities(suspList, from = "pubchemlite", adducts = "[M+H]+",
                             CCSParams = CCSParams)
# screen suspects and only keep fair matches (3% CCS tolerance)
IMSMatchParams <- getIMSMatchParams("CCS", window = 0.03, relative = TRUE)
fGroups <- screenSuspects(fGroups, suspects = suspList, adduct = "[M+H]+",
                          IMSMatchParams = IMSMatchParams, onlyHits = TRUE)

# perform feature annotation
mslists <- generateMSPeakLists(fGroups)
formulas <- generateFormulas(fGroups, mslists, "genform", adduct = "[M+H]+")
# skip compound annotation for IMS features with similar MS/MS data as their IMS precursor
compounds <- generateCompounds(fGroups, mslists, "metfrag", adduct = "[M+H]+", database = "pubchemlite",
                               minIMSSpecSim = 0.9)
# NOTE: assume that PubChemLite with CCS data is installed, so only mobility conversions needs to be applied
compounds <- assignMobilities(compounds, fGroups, adduct = "[M+H]+", CCSParams = CCSParams)
# filter out deviating candidates
compounds <- filter(compounds, IMSMatchParams = IMSMatchParams)

report(fGroups, mslists, formulas, compounds)
```

The [patRoonDataIMS] package contains example data and can be used to run this workflow. The data is already converted to various formats, therefore, the `convertMSFiles()` steps can be skipped when using this data. To use its data, simply assign the anaInfo variable:

```{r exIMSWorkflowData, eval=FALSE}
anaInfo <- patRoonDataIMS::exampleAnalysisInfo("positive") # positive IMS-HRMS data
```
