# Installation

`patRoon` depends on various other software tools to perform the non-target analysis workflow steps and to implement various other functionality. Most of these dependencies are automatically installed when you install the `patRoon` R package, however, some may need to be manually installed and/or configured.

> **_NOTE_**  It is highly recommended to perform installation steps in a 'clean' `R` session to avoid errors when installing or upgrading packages. As such it is recommended to close all open (R Studio) sessions and open a plain R console to perform the installation. 
 
## Automatic installation (Windows only)

An installation script is provided that automatically installs and configures all dependencies and finally installs `patRoon` itself. At this moment, this script only works with Microsoft Windows. You don't have to install anything else to use it, simply open `R` and execute these commands:

```{r eval=FALSE}
source("https://raw.githubusercontent.com/rickhelmus/patRoon/master/install_patRoon.R")
installPatRoon()
```

A simple text based wizard will start and asks you what to install and how to do it. You can re-run this installer at any time, for instance, if something went wrong or you want to install additional dependencies.

## Docker image

Docker images are provided to easily install a reproducible environment with `R`, `patRoon` and nearly all of its dependencies. This section assumes you have a basic understanding of [Docker] and have it installed. If not, please refer to the many guides available on the Internet. The Docker images of `patRoon` were originally only used for automated testing, however, since these contain a complete working environment of `patRoon` they are also suitable for using the software. They come with all external dependencies (except ProteoWizard), `R` dependencies and `MetFrag` libraries. Furthermore, the Docker image also contains [RStudio] server, which makes using `patRoon` even easier.

Below are some example shell commands on how to run the image.

```{bash, eval=FALSE}
# run an interactive R console session
docker run --rm -it patroonorg/patroonrs

# run a linux shell, from which R can be launched
docker run --rm -it patroonorg/patroonrs bash

# run rstudio server, accessible from localhost:8787
# login with rstudio/yourpasswordhere
docker run --rm -p 8787:8787 -u 0 -e PASSWORD=yourpasswordhere patroonorg/patroonrs /init

# same as above, but mount a local directory (~/myvolume) as local volume so it can be used for persistent storage
# please ensure that ~/myvolume exists!
docker run --rm -p 8787:8787 -u 0 -e PASSWORD=yourpasswordhere -v ~/myvolume:/home/rstudio/myvolume patroonorg/patroonrs /init
```

Note that the first two commands run as the default user `rstudio`, while the last two as `root`. The last commands launch [RStudio] server. You can access it by browsing to `localhost:8787` and logging in with user `rstudio` and the password defined by the `PASSWORD` variable from the command (`yourpasswordhere` in the above example). The last command also links a local volume in order to obtain persistence of files in the container's home directory. The Docker image is based on the excellent work from the [rocker project](https://www.rocker-project.org/). For more information on RStudio related options see their documentation for the [RStudio image].


## Manual installation

The manual installation is for users who don't use Windows or Docker, prefer to do a manual installation or simply want to know what happens behind the scenes. The manual installation consists of three phases:

1. Installing some prerequisite `R` packages
2. Install and configure (non-`R`) dependencies
3. Install `patRoon`

### R prerequisites

When installing `patRoon` Windows users have the option to install from a customized ([miniCRAN]) repository (`patRoonDeps`). This repository provides a central repository for `patRoon` and all its R packages. An advantage is that installation will be faster and you will not need to install [Rtools]. Note that you will need to have the latest `R` version installed in order to use this repository.

When you decide to use the `patRoonDeps` repository you can simply _skip_ this step. **Otherwise** (i.e. you will use regular repositories instead), the following will install the `R` dependencies.

```{r eval=FALSE}
install.packages(c("BiocManager", "remotes"))
BiocManager::install("CAMERA")

# needed for finding/grouping features with KPIC2
BiocManager::install("ropls")
remotes::install_github("rickhelmus/KPIC2") 

# needed for RAMClustR for componentization
remotes::install_github("cbroeckl/RAMClustR")

# needed for nontarget for componentization
remotes::install_github("blosloos/nontargetData")
remotes::install_github("blosloos/nontarget")

# needed for cliqueMS for componentization
remotes::install_github("rickhelmus/cliqueMS") # note: repository is a fork with minor bug fixes

# only needed for Bruker DataAnalysis integration
# NOTE: a fork is installed that fixes some issues with the latest R versions
remotes::install_github("BSchamberger/RDCOMClient")

# only when using the R interface (not recommended by default)
remotes::install_github("c-ruttkies/MetFragR/metfRag")

# needed for feature quality calculations with MetaClean
BiocManager::install(c("BiocStyle", "Rgraphviz")) 
remotes::install_github("KelseyChetnik/MetaClean")
```

Note that only the `CAMERA` installation is mandatory, the rest involves installation of _optional_ packages. If you are unsure which you need then you can always install the packages at a later stage.

### Other dependencies

Depending on which functionality is used, the following external dependencies may need to be installed:

Software                            | Remarks
----------------------------------- | -----------------------------------------------------
[Java JDK][JavaJDK]                 | **Mandatory** for e.g. plotting structures and using MetFrag.
[Rtools]                            | Necessary on Window and when `patRoon` is _not_ installed from `patRoonDeps`.
[ProteoWizard]                      | Needed for automatic data-pretreatment (e.g. data file conversion and centroiding, Bruker users may use DataAnalysis integration instead).
[OpenMS]                            | Recommended. Used for e.g. finding and grouping features.
[MetFrag CL][MetFragCL]             | Recommended. Used for annotation with MetFrag.
[MetFrag CompTox DB][CompTox-dl]    | Database files necessary for usage of the [CompTox] database with MetFrag. Note that a recent version  of MetFrag (>=2.4.5) is required. Note that the lists with additions for [smoking metadata][CompTox-smoke] and [wastewater metadata][CompTox-WW] are also supported.
[MetFrag PubChemLite DB][PCLite-dl] | Database file needed to use [PubChemLite][PCLite-paper] with MetFrag.
[SIRIUS]                            | For obtaining feature data and formula and/or compound annotation.
[BioTransformer]                    | For prediction of transformation products. See the [BioTransformer] page for installation details. If you have trouble compiling the jar file you can download it from [here](https://github.com/rickhelmus/patRoonDeps/raw/master/ext/biotransformer-3.0.0.jar).
[SAFD]                              | For finding features with [SAFD]. Please follow all the installation on the [SAFD webpage][SAFD].
[OpenBabel]                         | Used in some cases for suspect screening (e.g. to calculate molecular masses for suspects with only InChI information). Otherwise optional.
[pngquant]                          | Used to reduce size of HTML reports, definitely optional.

Most of these packages are optional and only needed if their algorithms are used during the workflow. If you are unsure which are needed, simply skip them for now and install them at a later stage.

After installation of these dependencies you may need to configure their filepaths (OpenMS and ProteoWizard are usually found automatically). To configure the file locations you should set some global package options with the `options()` R function, for instance:

```{r, eval=FALSE}
options(patRoon.path.pwiz = "C:/ProteoWizard") # location of ProteoWizard installation folder
options(patRoon.path.SIRIUS = "C:/sirius-win64-3.5.1") # location where SIRIUS was extracted
options(patRoon.path.OpenMS = "/usr/local/bin") # directory with the OpenMS binaries
options(patRoon.path.pngquant = "~/pngquant") # directory containing pngquant binary
options(patRoon.path.MetFragCL = "~/MetFrag2.4.5-CL.jar") # full location to the jar file
options(patRoon.path.MetFragCompTox = "C:/CompTox_17March2019_SelectMetaData.csv") # full location to desired CompTox CSV file
options(patRoon.path.MetFragPubChemLite = "~/PubChemLite_14Jan2020_tier0.csv") # full location to desired PubChemLite CSV file
options(patRoon.path.BioTransformer = "~/biotransformer/biotransformer-3.0.1.jar")
options(patRoon.path.obabel = "C:/Program Files/OpenBabel-3.0.0") # directory with OpenBabel binaries
```

These commands have to be executed everytime you start a new R session (e.g. as part of your script). However, it is probably easier to add them to your `~/.Rprofile` file so that they are executed automatically when you start R. If you don't have this file yet you can simply create it yourself (for more information see e.g. [this SO answer](https://stackoverflow.com/a/46819910)).

### patRoon installation

Finally, it is time to install `patRoon` itself. As mentioned before, Windows users (who have the latest `R` version) can install `patRoon` and all its package dependencies from the `patRoonDeps` repository:

```{r eval=FALSE}
# install from patRoonDeps (only Windows with latest R version)
install.packages("patRoon", repos = "https://rickhelmus.github.io/patRoonDeps/", type = "binary")
```

**Otherwise**, installation occurs directly from GitHub:

```{r eval=FALSE}
remotes::install_github("rickhelmus/patRoon")
```

Optional example data is installed via GitHub:

```{r eval=FALSE}
# optional example data
remotes::install_github("rickhelmus/patRoonData")
```


Afterwards, you can run the `verifyDependencies()` function to see if `patRoon` can find all its dependencies (you may need to restart R beforehand):

```{r eval=FALSE}
patRoon::verifyDependencies()
```

```{r child=file.path(vignDir, "shared", "_refs.Rmd")}
```
