# Transformation product screening {#TPs}

This chapter describes the various functionality for screening of _transformation products_ (TPs), which are introduced since `patRoon` 2.0. Screening for TPs, i.e. chemicals that are formed from a _parent_ chemical by e.g. chemical or biological processes, has broad applications. For this reason, the TP screening related functionality is designed to be flexible, thus allowing one to use a workflow that is best suited for a particular study.

Regardless, the TP screening workflow in `patRoon` can be roughly summarized as follows:

```{r TPWorkflow,echo=FALSE,out.width="100%"}
plotGV("
digraph Workflow {
  graph [ rankdir = LR ]
  node [ shape = box,
         fixedsize = true,
         width = 2.2,
         height = 1,
         fontsize = 18,
         fillcolor = darkseagreen1,
         style = filled ]

    'Parent screening' -> 'Obtaining TPs' -> 'TP screening' -> 'Linking parent/TPs'
}", height = 90, width = 500)
```

* **Parent screening** During this step a common `patRoon` workflow is used to screen for the parent chemicals of interest. This could be a full non-target analysis with compound annotation or a relative simple suspect or target screening.
* **Obtaining TPs** Data is obtained of potential TPs for the parents of interest. The TPs may originate from a library or predicted _in-silico_. Note that in some workflows this step is omitted (discussed later).
* **TP screening** A suspect screening is performed to find the TPs in the analysis data. Some workflows skip this step.
* **Linking parents and TPs** In the step the parent features are linked with the TP features. Several post-processing functionality exists to improve and prioritize the data.

The next sections will outline more details on these steps are performed and configured. The [last section](#TPsExamples) in this chapter outlines several example workflows.

> **_NOTE_** The [newProject tool](#newProject) can be used to easily generate a workflow with transformation product screening.

## Obtaining transformation product data {#genTPs}

The `generateTPs` function is used to obtain TPs for a particular set of parents. Like other workflow generator functions (`findFeatures`, `generateCompounds`), several algorithms are available that do the actual work.


Algorithm            | Usage                                             | Input                        | Output
-------------------- | ------------------------------------------------- | ---------------------------- | ------------------------------------------
[BioTransformer]     | `generateTPs(algorithm = "biotransformer", ...)`  | Parents                      | TPs structural information
[CTS]                | `generateTPs(algorithm = "cts", ...)`             | Parents                      | TPs with structural information
Library              | `generateTPs(algorithm = "library", ...)`         | Parents (optional), Library ([PubChem][PubChemLiteTR] or custom) | TPs with structural information
Formula library      | `generateTPs(algorithm = "library_formula", ...)` | Parents                      | TPs with formula information
Metabolic logic      | `generateTPs(algorithm = "logic", ...)`           | Feature groups | TPs from _m/z_ differences from pre-defined elemental transformations (based on @Scholle2015).
Formula Annotations  | `generateTPs(algorithm = "ann_form", ...)`        | Parents, formula annotations | Prioritized TPs from annotation candidates (based on @Helmus2025).
Compound Annotations | `generateTPs(algorithm = "ann_comp", ...)`        | Parents, compound annotaions | Prioritized TPs from  annotation candidates (based on @Helmus2025).

For most workflows the `biotransformer`, `cts` and `library` algorithms are a good starting point. They are fairly straightforward to use and output TPs with full structural information. The `library_formula` algorithm may be a suitable alternative if only parents with formula information are available. The `ann_comp` and `ann_form` algorithms are meant to elucidate completely unknown TPs, i.e. those that are not predicted or found in literature, and are discussed [further below](#TPsAnn). Finally, the `logic` algorithm is primarily meant if no or very little information on possible parents and/or TPs is available.

The parent information that is needed for most algorithms is taken from one of the following:

1. The data in a suspect list (follows the same format as [suspect screening](#suspscr))
2. The data from suspects that were matched to feature groups (e.g. obtained with `screenSuspects`, see [suspect screening](#suspscr))
3. A `compounds` object obtained with [compound annotation](#compounds) (only `biotransformer`, `cts` and `library`)

The second option is often used in most workflows. The use of compound annotations as parent input could be considered if the parents are not  known. However, care must be taken since _all_ the candidates are used, and it is highly recommend to filter the object in advance with e.g. the `topMost` filter. For `library` and `library_formula`, the parent input is optional: if no parents are specified (`parents=NULL`) then TP data for _all_ parents in the database is used.

An overview of common arguments for TP generation is listed below.

Argument                      | Algorithm(s)      | Remarks
----------------------------- | ----------------- | --------------------------------------------------------
`parents`                     | `biotransformer`, `cts`, `library`, `library_formula`, `ann_form`, `ann_comp` | The input parents.
`fGroups`                     | `logic`           | The input feature groups to calculate TPs for.
`type`                        | `biotransformer`  | The prediction type: `"env"`, `"ecbased"`, `"cyp450"`, `"phaseII"`, `"hgut"`, `"superbio"`, `"allHuman"`. See [BioTransformer] for more details.
`transLibrary`                | `cts`             | The transformation library that should be used: `"hydrolysis"`, `"abiotic_reduction"`, `"photolysis_unranked"`, `"photolysis_ranked"`, `"mammalian_metabolism"`, `"combined_abioticreduction_hydrolysis"`, `"combined_photolysis_abiotic_hydrolysis"`, `"pfas_environmental"`, `"pfas_metabolism"`. See [cts] for more details.
`TPLibrary`                   | `library`, `library_formula` | [Custom TP library](#TPsCustom).
`transformations`             | `logic`           | [Custom TP transformation rules](#TPsCustom).
`generations`                 | `biotransformer`, `cts`, `library`, `library_formula` | The number of transformation generations to consider.
`adduct`                      | `logic`           | The assumed adduct of the parents (e.g. `"[M+H]+"`). Not needed when [adduct annotations](#incorpAdductIso) are available.
`formulas`/`compounds`        | `ann_form`/`ann_comp` | The input formula/compound annotations to extract TPs from.
`TPsRef`, `fGroupsComps`      | `ann_comp`        | The reference TPs and feature groups to use for candidate ranking. See [below](#TPsAnn) for more details.
`TPStructParams`              | `biotransformer`, `cts`, `library`, `ann_comp` | Other advanced parameters, e.g. to calculate structural similarities. See `?getDefTPStructParams` for more details.

Some examples on how to generate TPs are shown below:

```{r eval=FALSE}
# predict environmental TPs with BioTransformer for all parents in a suspect list
TPsBT <- generateTPs("biotransformer", parents = patRoonData::suspectsPos,
                     type = "env")
# obtain all TPs from the default library
TPsLib <- generateTPs("library")
# get TPs for the parents matched in a suspect screening
TPsLib <- generateTPs("library", parents = fGroupsScr)

# calculate TPs for all feature groups
TPsLogic <- generateTPs("logic", fGroups, adduct = "[M+H]+")

# use formula annotations to obtain TPs
TPsAnnForm <- generateTPs("ann_form", formulas = formulas)

# use compound annotations to obtain TPs and provide additional suspect and feature data for candidate ranking
TPsAnnComp <- generateTPs("ann_comp", compounds = compounds, TPsRef = TPsBT, fGroupsComps = fGroups)
```

### (Custom) Libraries and transformations {#TPsCustom}

By default the `library` and `logic` algorithms use data that is installed with `patRoon` (based on [PubChem transformations][PubChemLiteTR] and @Scholle2015, respectively). However, it is also possible to use custom data. For the `library_formula` no default library is provided, however, these can easily be generated as is discussed at the end of the section. 

To use a custom TP structure library for the `library` algorithm a simple `data.frame` is needed with the names, SMILES and optionally `log P` values for the parents and TPs. The `log P` values are used for prediction of the retention time direction of a TP compared to its parent, as is [discussed later](#TPCompProc). The following small library has two TPs for benzotriazole and one for DEET:

```{r TPsCustomDB,eval=TRUE}
myTPLib <- data.frame(parent_name = c("1H-Benzotriazole", "1H-Benzotriazole", "DEET"),
                      parent_SMILES = c("C1=CC2=NNN=C2C=C1", "C1=CC2=NNN=C2C=C1", "CCN(CC)C(=O)C1=CC=CC(=C1)C"),
                      TP_name = c("1-Methylbenzotriazole", "1-Hydroxybenzotriazole", "N-ethyl-m-toluamide"),
                      TP_SMILES = c("CN1C2=CC=CC=C2N=N1", "C1=CC=C2C(=C1)N=NN2O", "CCNC(=O)C1=CC=CC(=C1)C"))
myTPLib
```

To use this library, simply pass it to the `TPLibrary` argument:

```{r eval=FALSE}
TPs <- generateTPs("library", TPLibrary = myTPLib)
```

For `library_formula` the library follows the same format. However, here the formula should be specified instead of the `SMILES` with the `parent_formula` and `TP_formula` columns (although it is still allowed to only specify SMILES, in which case the formulae are calculated from the SMILES).

For the `logic` algorithm a table with custom transformation rules can be specified for TP calculations:

```{r TPCustomLogic,eval=TRUE}
myTrans <- data.frame(transformation = c("hydroxylation", "demethylation"),
                      add = c("O", ""),
                      sub = c("", "CH2"),
                      retDir = c(-1, -1))
myTrans
```

The `add` and `sub` columns are used to denote the elements that are added or subtracted by the reaction. These are used to calculate mass differences between parents and TPs. The `retDir` column is used to indicate the retention time direction of the parent compared to the TP: `-1` (elutes before parent), `1` (elutes after parent) or `0` (similar or unknown). This is [discussed later](#TPCompProc) how this data can be used to filter TP candidates. The custom rules can be used by passing them to the `transformations` argument:

```{r eval=FALSE}
TPs <- generateTPs("logic", fGroups, adduct = "[M+H]+", transformations = myTrans)
```

The `genFormulaTPLibrary()` utility function can be used to automatically generate TP libraries suitable for the `library_formula` algorithm. The transformation rules to calculate TPs are specified in the same format as used by the `logic` algorithm.

```{r TPFormLib,eval=FALSE}
myTPFormLib <- genFormulaTPLibrary(parents = patRoonData::suspectsPos, transformations = myTrans)
# also calculate second generation TPs (TPs of TPs)
myTPFormLib2 <- genFormulaTPLibrary(parents = patRoonData::suspectsPos, transformations = myTrans,
                                    generations = 2)

# Use library
TPs <- generateTPs("library_formula", TPLibrary = myTPFormLib)
```

Compared to the `logic` algorithm, the `library_formula` algorithm is more (and only) suitable for suspect/target screening workflows, allows multiple transformation generations and allows better customization through manually adding/removing TPs from the library prior to passing it to `generateTPs()`.

### TPs from feature annotation candidates {#TPsAnn}

The `ann_form` and `ann_comp` algorithms are intended to provide a thorough screening and elucidation of TPs that are otherwise difficult to find with other algorithms. These algorithms assume that TPs of interest can be revealed from a thorough feature annotation workflow based on formulae (`ann_form`) or compounds (`ann_comp`). The annotation candidates are prioritized and ranked by the _TP Score_, which is calculated from properties such as the fit of the candidate structure or formula into the parent (or vice versa) and similarity to suspects (i.e. TPs obtained by other algorithms). For more details see @Helmus2025, `?generateTPsAnnForm` and `?generateTPsAnnComp`.

To provide a thorough screening with `ann_comp`, often a large compound database such as PubChem is used for compound annotation. This typically results in tens of thousands of candidates for each parent. Hence, obtaining the compound annotations and generating TPs with `ann_comp` is computationally intensive and can take multiple hours. Afterwards, the `topMost` filter should be used to reduce the number of candidates to a manageable size (see the next section).


### Processing data {#TPsProc}

Similar to other workflow data, several generic functions are available to inspect the TP data:

Generic                            | Remarks                                                             
-----------------------------------|----------------------------------------------------------------------
`length()`                         | Returns the total number of transformation products
`names()`                          | Returns the names of the parents
`parents()`                        | Returns a table with information about the parents
`products()`                       | Returns a `list` with for each parent a table with TPs
`as.data.table()`, `as.data.frame` | Convert all the object information into a `data.table`/`data.frame`
`"[["` / `"$"` operators           | Extract TP information for a specified parent

Some examples:

```{r TPsProcInsp,eval=runData}
# just show a few columns in this example, there are many more!
# note: the double dot syntax (..cols) is necessary since the data is stored as data.tables
cols <- c("name", "formula", "InChIKey")
parents(TPs)[1:5, ..cols]
TPs[["DEET"]][, ..cols]
TPs[[2]][, ..cols]
as.data.table(TPs)[1:5, 1:3]
```

In addition, the following generic functions are available to modify or convert the object data:

Generic                 | Remarks
----------------------- | ---------------------------------------------------------------------
`"["` operator          | Subset this object on given parents
`filter`                | Filters this object (available functionality depends on TP generation algorithm)
`convertToSuspects`     | Generates a suspect list of all TPs (and optionally parents) that is suitable for `screenSuspects`
`convertToMFDB`         | Generates a [MetFrag] database for all TPs (and optionally parents, only for TPs with structural information)
`plotGraph`             | Generates an interactive plot to explore transformation hierarchies (only for TPs with structural or formula information)
`plotVenn`, `plotUpSet` | Compare results between different algorithms with Venn/UpSet diagrams (only for TPs with structural information)
`consensus`             | Combine results from different algorithms (only for TPs with structural information). See the [algorithm consensus](#consensus) section.


The `convertToMFDB` function is especially handy with predicted TPs, as it allows generating a compound database for TPs that may not be available in commonly used databases. This is further demonstrated in the [first example](#TPsEx1).

```{r TPsProcMod,eval=FALSE}
TPs2 <- TPs[1:10] # only keep results for first ten parents

# only keep TPs with likely/probably likelihood (specific property for CTS algorithm)
TPsF <- filter(TPs, properties = list(likelihood = c("LIKELY", "PROBABLE")))

# remove transformation products that are isomers to their parent or sibling TPs
# may simplify data as these are often difficult to identify
TPsF <- filter(TPs, removeParentIsomers = TRUE, removeTPIsomers = TRUE)

# remove duplicate transformation products from each parent
# these can occur if different pathways yield the same TPs
TPsF <- filter(TPs, removeDuplicates = TRUE)

# only keep TPs that have a structural similarity to their parent of >= 0.5
# (set TPStructParams=getDefTPStructParams(calcSims=TRUE) when executing generateTPs())
TPsF <- filter(TPs, minSimilarity = 0.5)

# only keep TPs with a TP Score >= 0.5 and the highest 25 TPs per parent
# see ?transformationProductsAnnComp for specific filters
TPsAnnCompF <- filter(TPsAnnComp, minTPScore = 0.5, topMost = 25)

# do a suspect screening for all TPs and their parents
# this is often part of a workflow and is discussed further in the next section.
suspects <- convertToSuspects(TPs, includeParents = TRUE)
fGroupsScr <- screenSuspects(fGroups, suspects, onlyHits = TRUE)

# use the TP data for a specialized MetFrag database
convertToMFDB(TPs, "TP-database.csv", includeParents = FALSE)
compoundsTPs <- generateCompounds(fGroups, mslists, "metfrag", database = "csv",
                                  extraOpts = list(LocalDatabasePath = "TP-database.csv"))
```

```{r TPsPlots,eval=runData}
plotVenn(TPsLib, TPsBT, labels = c("lib", "BT"))
plotGraph(TPsBT, which = "Triazophos") # hierarchy for Triazophos parent
```


## Linking parent and transformation product features

This section discusses one of the most important steps in a TP screening workflow, which is to link feature groups of parents with those of candidate transformation products. During this step, _TP components_ are made, where each component consist of one or more feature groups of detected TPs for a particular parent. Note that componentization was [already introduced before](#componentization), but for very different algorithms. However, the data format for TP componentization is quite similar. After componentization, several filters are available to clean and prioritize the data. These can even allow workflows without obtaining potential TP candidates in advance, which is discussed in the last subsection.

### Componentization

Like [other algorithms](#components), the `generateComponents` generic function is used to generate TP components, by setting the `algorithm` parameter to `"tp"`.

The following arguments are of importance:

Argument        | Remarks
--------------- | --------------------------------------------------------------
`fGroups`       | The input feature groups for the _parents_
`fGroupsTPs`    | The input feature groups for the _TPs_
`ignoreParents` | Set to `TRUE` to ignore feature groups in `fGroupsTPs` that also occur in `fGroups`
`TPs`           | The input transformation products, ie as generated by `generateTPs()`
`MSPeakLists`, `formulas`, `compounds` | Annotation objects used for similarity calculation between the parent and its TPs
`minRTDiff`     | The minimum retention time difference (seconds) of a TP for it to be considered to elute differently than its parent.

#### Feature group input {#TPsFGroups}

The `fGroups`, `fGroupsTPs` and `ignoreParents` arguments are used by the componentization algorithm to identify which feature groups can be considered as parents and which as TPs. Three scenarios are possible:

1. `fGroups=fGroupsTPs` and `ignoreParents=FALSE`: in this case no distinction is made, and all feature groups are considered a parent or TP (default if `fGroupsTPs` is not specified).
2. `fGroups` and `fGroupsTPs` contain different subsets of the _same_ `featureGroups` object and `ignoreParents=FALSE`:  only the feature groups in `fGroups`/`fGroupsTPs` are considered as parents/TPs.
3. As above, but with `ignoreParents=TRUE`: the same distinction is made as above, but any feature groups in `fGroupsTPs` are ignored for TP candidate selection if also present in `fGroups`.

The first scenario is often used if it is unknown which feature groups may be parents or which are TPs. Furthermore, this scenario may also be used if the dataset is sufficiently simple, for instance, because a suspect screening with the results from `convertToSuspects` (discussed in the previous section) would reliably discriminate between parents and TPs. A workflow with the first scenario is demonstrated in the [second example](#TPsEx2).

In all other cases it is recommended to use either the second or third scenario, since making a prior distinction between parent and TP feature groups greatly simplifies the dataset and reduces false positives. A relative simple example where this can be used is when there are two sample groups: before and after treatment.

```{r eval=FALSE}
componTP <- generateComponents(algorithm = "tp",
                               fGroups = fGroups[ni = treatment == "before"],
                               fGroupsTPs = fGroups[ni = treatment == "after"])
```

In this example, only those feature groups with features present in the "before" treatment group are considered as parents, and those in "after" may be considered as a TP (see [this section](#useMeta) on how to group sample analyses with metadata). Since it is likely that there will be some overlap in feature groups between both sample groups, the `ignoreParents` flag can be used to not consider any of the overlap for TP assignments:

```{r eval=FALSE}
componTP <- generateComponents(algorithm = "tp",
                               fGroups = fGroups[ni = treatment == "before"],
                               fGroupsTPs = fGroups[ni = treatment == "after"],
                               ignoreParents = TRUE)
```

More sophisticates ways are of course possible to provide an upfront distinction between parent/TP feature groups. In the [fourth example](#TPsEx4) a workflow is demonstrated where fold changes are used.

> **_NOTE_** The feature groups specified for `fGroups`/`fGroupsTPs` _must_ always originate from the same `featureGroups` object.

If TPs were generated with an algorithm that requires parent input (see [Obtaining TPs](#genTPs)), then it is often mandatory that a suspect screening of parents and TPs is performed prior to componentization. This is necessary for the componentization algorithm to map the feature groups that belong to a particular parent or TP. Note that this step should _not_ be performed for the `ann_form` and `ann_comp` algorithms, as these algorithms already provide the necessary mappings. The `convertToSuspects` function is used to prepare the suspect list:

```{r eval=FALSE}
# perform suspect screening
# NOTE: set includeParents to TRUE since both the parents and TPs should be screened
# NOTE: for the ann_form and ann_comp algorithms no suspect screening is necessary
suspects <- convertToSuspects(TPs, includeParents = TRUE)
fGroupsScr <- screenSuspects(fGroups, suspects, onlyHits = TRUE)

# do the componentization
# a similar distinction between fGroups/fGroupsScr as discussed above can of course also be done
componTP <- generateComponents(fGroups = fGroupsScr, ...)
```

If a parent screening was already performed in advance, for instance when the input parents to `generateTPs` are screening results, the screening results for parents and TPs can also be combined. The [second example](#TPsEx2) demonstrates this.

Note that in the case a parent suspect is matched to multiple feature groups, a component is made for each match. Similarly, if multiple feature groups match to the same TP suspect, all of them will be incorporated in the component.

When TPs were generated with the `logic` algorithm a suspect screening must also be carried out in advance. However, in this case it is not necessary to include the parents (since each parent equals a feature group no mapping is necessary). The `onlyHits` variable to `screenSuspects` must not be set in order to keep the parents.

```{r eval=FALSE}
# only screen for TPs
suspects <- convertToSuspects(TPs, includeParents = FALSE)
# but keep all other feature groups as these may be parents
fGroupsScr <- screenSuspects(fGroups, suspects, onlyHits = FALSE)

# do the componentization...
```

#### Annotation similarity calculation

If additional annotation data for parents and TPs is given to the componentization algorithm, it will be used to calculate various similarity metrics. Often, the chemical structure for a transformation product is similar to that of its parent. Hence, there is a good chance that a parent and its TPs also share similar MS/MS data.

Firstly, if MS peak lists are provided, then the [spectrum similarity](#specSim) is calculated between each parent and its potential TP candidates. This is performed with all the three different alignment shifts (see the [spectrum similarity section](#specSim) for more details).

In case `formulas` and/or `compounds` objects are given as input to `generateComponents()`, then a parent/TP comparison is made by counting the number of fragments and neutral losses that they share (based on the formulae assigned to the MS/MS fragments). The counts are calculated in two different ways:

1. from the matches between the parent and the TP candidate
2. from the matches between the parent and _all_ the annotation candidates in the input `formulas`/`compounds` object for the TP feature group 

The latter is mainly used (and only available) in workflows [where componentization is performed without previously generated TPs](#TPsOmit). To improve the usefulness of the total similarity metric, it is highly recommend to pre-treat the annotation objects with e.g. the `topMost` filter. Both calculation methods pool the data from the input `formulas` and `compounds` and only count unique fragment/neutral loss matches.

### Processing data {#TPCompProc}

The output of TP componentization is an object of the `componentsTPs` class. This _derives_ from the 'regular' `components` class, therefore, all the data processing functionality described [before](#processing) (extraction, subsetting, filtering etc) are also valid for TP components.

For the `as.data.table()` method function (and `as.data.frame()`) the `candidates` argument can be used to specify if individual candidates for each feature group in the component should be included in the output:

```{r TPCompProcADT,eval=runData}
# only output feature group data
as.data.table(componTP, candidates = FALSE)[name == "CMP2", .(name, parent_name, group)]
# also include the candidates for each feature group
as.data.table(componTP, candidates = TRUE)[name == "CMP2", .(name, parent_name, group, candidate_name, formula)]
```

Several additional filters are available to prioritize the data:

Filter        | Remarks
------------- | -----------------------------------
`retDirMatch` | If `TRUE` only keep TPs with an expected chromatographic retention direction compared to the parent.
`minSpecSim`, `minSpecPrec`, `minSpecSimBoth` | The minimum spectrum similarity between the parent and TP. Calculated with no, `"precursor"` and `"both"` alignment shifting (see [spectrum similarity](#specSim)).
`minFragMatches`, `minNLMatches` | Minimum number of formula fragment/neutral loss matches between parent and TP (discussed in previous section).
`minTotFragMatches`, `minTotNLMatches` | As above, but for total matches.
`formulas`    | A `formulas` object used to further verify candidate TPs that were generated by the `logic` algorithm.

The `retDirMatch` filter compares the expected and observed _retention time direction_ of a TP in order to decide if it should be kept. The direction is a value of either `-1` (TP elutes before parent), `+1` (TP elutes after parent) or `0` (TP elutes very close to the parent or its direction is unknown). The directions are taken from the [generated transformation products](#genTPs). In most cases the log P values are compared of a TP and its parent. Here, it is assumed that lower log P values result in earlier elution (i.e. typical with reversed phase LC). For the `logic` algorithm the retention time direction is taken from the transformation rules table. Note that specifying a large enough value for the `minRTDiff` argument to `generateComponents` is important to ensure that some tolerance exists while comparing retention time directions of parent and TPs. Furthermore, the `TPStructParams` argument to `generateTPs()` can be used to tweak the calculation of expected retention time directions from log P values (see `?getDefTPStructParams`). This filter does nothing if either the observed or expected direction is zero.

When TP data was generated with the `logic` algorithm it is recommended to use the `formulas` filter. This filter uses formula annotations to verify that (1) a parent feature group contains the elements that are subtracted during the transformation and (2) the TP feature group contains the elements that were added during the transformation. Since the 'right' candidate formula is most likely not yet known, this filter looks at _all_ candidates. Therefore, it is recommended to filter the `formulas` object, for instance, with the `topMost` filter.

Finally, the `plotGraph()` method function that was introduced exploring transformation hierarchies for structure TPs, can also incorporate componentization results to simplify the plot and mark TP hits:

```{r plotGraphCompTP,eval=runData}
plotGraph(TPsBT, which = "Atrazine", components = componTP)
```

### Omitting transformation product input {#TPsOmit}

The `TPs` argument to `generateComponents` can also be omitted (i.e. `TPs=NULL`). In this case every feature group from `fGroupTPs` is considered to be a potential TP for the potential parents specified for `fGroups`. An advantage is that the screening workflow is not limited to any known TPs or transformations. However, such a workflow has high demands on prioritiation steps before and after the componentization to rule out the many false positives that may occur.

When no transformation data is supplied it is crucial to make [a prior distinction](#TPsFGroups) between parent and TP feature groups. Afterwards, the MS/MS spectral and other annotation similarity filters mentioned in the previous section may helpful to further prioritize data.

The [fourth example](#TPsEx4) demonstrates such a workflow.

### Reporting TP components

The TP components can be reported with the `report` function. This is done by setting the `components` function argument (i.e. equally to all other component types). The results will be displayed with a customized format that allows easy exploring of each parent with its TPs. In addition, the `TPs` argument can be set to include additional data such as transformation pathways.

```{r eval=FALSE}
report(fGroups, components = componTP, TPs = TPs)
```

## Example workflows {#TPsExamples}

The next subsections demonstrate several approaches to perform a TP screening workflow with `patRoon`. In all examples it is assumed that feature groups were already obtained (with the `findFeatures` and `groupFeatures` functions) and stored in the `fGroups` variable.

The workflows with `patRoon` are designed to be flexible, and the examples here are primarily meant to implement your own workflow. Furthermore, some of the techniques used in the examples can also be combined. For instance, the [Fold change](#FCCalc) classification and MS/MS similarity filters applied in the [fourth example](#TPsEx4) could also be applied to any of the other examples.

### Screen predicted TPs for targets {#TPsEx1}

The first example is a simple workflow where TPs are predicted for a set of given parents with [BioTransformer] and subsequently screened. A [MetFrag] compound database is generated and used for annotation.

```{r eval=FALSE}
# predict TPs for a fixed list of parents
TPs <- generateTPs("biotransformer", parents = patRoonData::suspectsPos)

# screen for the TPs
suspectsTPs <- convertToSuspects(TPs, includeParents = FALSE)
fGroupsTPs <- screenSuspects(fGroups, suspectsTPs, adduct = "[M+H]+", onlyHits = TRUE)

# perform annotation of TPs
mslistsTPs <- generateMSPeakLists(fGroupsTPs)
convertToMFDB(TPs, "TP-database.csv", includeParents = FALSE) # generate MetFrag database
compoundsTPs <- generateCompounds(fGroupsTPs, mslistsTPs, "metfrag", adduct = "[M+H]+", database = "csv",
                                  extraOpts = list(LocalDatabasePath = "TP-database.csv"))
```

### Screening TPs from a library for suspects {#TPsEx2}

In this example TPs of interest are obtained for the parents that surfaced from of a suspect screening. The steps of this workflow are:

1. Suspect screening parents.
2. Obtain TPs for the suspect hits from a library.
3. A second suspect screening is performed for TPs and the original parent screening results are amended. Note that the parent data is needed for componentization.
4. Both parents and TPs are annotated using a database generated from their chemical structures.
5. Some prioritization is performed by
    a. Only keeping candidate structures for which _in-silico_ fragmentation resulted in at least one annotated MS/MS peak.
    b. Only keeping suspect hits with an estimated identification level of 3 or better.
6. The TP components are made and only feature groups with parent/TP assignments are kept.
7. All results are reported.

```{r eval=FALSE}
# step 1
fGroupsScr <- screenSuspects(fGroups, patRoonData::suspectsPos, adduct = "[M+H]+")
# step 2
TPs <- generateTPs("library", parents = fGroupsScr)

# step 3
suspects <- convertToSuspects(TPs)
fGroupsScr <- screenSuspects(fGroupsScr, suspects, adduct = "[M+H]+", onlyHits = TRUE, amend = TRUE)

# step 4
mslistsScr <- generateMSPeakLists(fGroupsScr)
convertToMFDB(TPs, "TP-database.csv", includeParents = TRUE)
compoundsScr <- generateCompounds(fGroupsScr, mslistsScr, "metfrag", adduct = "[M+H]+", database = "csv",
                                  extraOpts = list(LocalDatabasePath = "TP-database.csv"))

# step 5a
compoundsScr <- filter(compoundsScr, minExplainedPeaks = 1)

# step 5b
fGroupsScrAnn <- estimateIDConfidence(fGroupsScr, MSPeakLists = mslistsScr,
                                      compounds = compoundsScr)
fGroupsScrAnn <- filter(fGroupsScrAnn, maxLevel = 3, onlyHits = TRUE)

# step 6
componTP <- generateComponents(fGroupsScrAnn, "tp", TPs = TPs, MSPeakLists = mslistsScr,
                               compounds = compoundsScr)
fGroupsScrAnn <- fGroupsScrAnn[results = componTP]

# step 7
report(fGroupsScrAnn, MSPeakLists = mslistsScr, compounds = compoundsScr,
       components = componTP, TPs = TPs)
```

### Non-target screening of predicted TPs {#TPsEx3}

This example uses metabolic logic to calculate possible TPs for all feature groups from a complete non-target screening. This example demonstrates how a workflow can be performed when little is known about the identity of the parents. The steps of this workflow are:

1. Formula annotations are performed for all feature groups.
2. These results are then limited to the top 5 candidates, and only feature groups with annotations are kept.
3. The TPs are calculated for all remaining feature groups.
4. A suspect screening is performed to find the TPs. Unlike the previous example feature groups without hits are kept ([discussed here](#TPsFGroups)).
5. The components are generated
6. The components are filtered:
    a. The TPs must follow an expected [retention time direction](#TPCompProc)
    b. The parent/TPs should have at least one candidate formula that fits with the transformation.
7. Only feature groups are kept with parent/TP assignments and all results are reported.

```{r eval=FALSE}
# steps 1-2
mslists <- generateMSPeakLists(fGroups)
formulas <- generateFormulas(fGroups, mslists, "genform", adduct = "[M+H]+")
formulas <- filter(formulas, topMost = 5)
fGroups <- fGroups[results = formulas]

# step 3
TPs <- generateTPs("logic", fGroups = fGroups, adduct = "[M+H]+")

# step 4
suspects <- convertToSuspects(TPs)
fGroupsScr <- screenSuspects(fGroups, suspects, adduct = "[M+H]+", onlyHits = FALSE)

# step 5
componTP <- generateComponents(fGroupsScr, "tp", TPs = TPs, MSPeakLists = mslists, formulas = formulas)

# step 6
componTP <- filter(componTP, retDirMatch = TRUE, formulas = formulas)

# step 7
fGroupsScr <- fGroupsScr[results = componTP]
report(fGroupsScr, MSPeakLists = mslists, formulas = formulas, components = componTP)
```

### Non-target screening of TPs by annotation similarities {#TPsEx4}

This example shows a workflow where no TP data from a prediction or library is used. Instead, this workflow relies on statistics and MS/MS data to find feature groups which may potentially have a parent - TP relationship. The workflow is similar to that of the previous example. The steps of this workflow are:

1. [Fold changes](#FCCalc) (FC) between two sample groups are calculated to classify which feature groups are decreasing (i.e. parents) or increasing (i.e. TPs).
2. Feature groups without classification are removed.
3. Formula annotations are performed like the previous example.
4. The componentization is performed and the FC classifications are used to specify which feature groups are to be considered parents or TPs.
5. Only TPs are kept that show a high MS/MS spectral similarity and share at least one fragment with their parent.
6. Only feature groups are kept with parent/TP assignments and all results are reported.

```{r eval=FALSE}
# step 1
tab <- as.data.table(fGroups, FCParams = getFCParams(c("before", "after")))
groupsParents <- tab[classification == "decrease"]$group
groupsTPs <- tab[classification == "increase"]$group

# step 2
fGroups <- fGroups[, union(groupsParents, groupsTPs)]

# step 3
mslists <- generateMSPeakLists(fGroups)
formulas <- generateFormulas(fGroups, mslists, "genform", adduct = "[M+H]+")
formulas <- filter(formulas, topMost = 5)
fGroups <- fGroups[results = formulas]

# step 4
componTP <- generateComponents(algorithm = "tp",
                               fGroups = fGroups[, groupsParents],
                               fGroupsTPs = fGroups[, groupsTPs],
                               MSPeakLists = mslists, formulas = formulas)

# step 5
componTP <- filter(componTP, minSpecSimBoth = 0.75, minTotFragMatches = 1)

# step 6
fGroups <- fGroups[results = componTP]
report(fGroups, MSPeakLists = mslists, formulas = formulas, components = componTP)
```
